---
title: "Bayesian modelling"
author: "Léo Belzile"
subtitle: "Monte Carlo methods and Markov chains"
date: today
date-format: "[Last compiled] dddd MMM D, YYYY"
eval: true
echo: true
cache: true
width: 1200
height: 900
bibliography: MATH80601A.bib
format:
  revealjs:
    slide-number: true
    html-math-method: mathjax
    preview-links: auto
    theme: [simple, hecmontreal.scss]
    title-slide-attributes:
      data-background-color: "#ff585d"
    logo: "fig/logo_hec_montreal_bleu_web.png"
---


```{r}
#| include: false
#| eval: true
#| echo: false
hecbleu <- c("#002855")
fcols <- c(gris = "#888b8d",
           bleu = "#0072ce",
           aqua = "#00aec7",
           vert = "#26d07c",
           rouge = "#ff585d",
           rose = "#eb6fbd",
           jaune = "#f3d03e")
pcols <- c(gris = "#d9d9d6",
           bleu = "#92c1e9",
           agua = "#88dbdf",
           vert = "#8fe2b0",
           rouge = "#ffb1bb",
           rose = "#eab8e4",
           jaune = "#f2f0a1")
library(ggplot2)
theme_set(theme_classic())
library(patchwork)
knitr::opts_chunk$set(fig.retina = 3, collapse = TRUE)
options(digits = 3, width = 75)
```

## Bayesian inference beyond conjugate models

How to circumvent the problem of intractable posteriors?

- simulation-based methods: accept-reject, Markov chain Monte Carlo, particle filters, etc.
- deterministic methods: (integrated nested) Laplace approximations, variational Bayes, expectation propagation, etc.

We focus on Monte Carlo methods.


## Simulation algorithms: inversion method

If $F$ is an absolutely continuous distribution function, then 
$$F(X) \sim \mathsf{unif}(0,1).$$
The inversion method consists in applying the quantile function $F^{-1}$ to $U \sim \mathsf{unif}(0,1)$, viz. $$F^{-1}(U) \sim X.$$



## Inversion method for truncated distributions

Consider a random variable $Y$ with distribution function $F$.

If $X$ follows the same distribution as $Y$, but restricted over the interval $[a,b]$, then
$$\Pr(X \leq x) = \frac{F(x) - F(a)}{F(b)-F(a)}, \qquad a \leq x \leq b,$$

Therefore, $$F^{-1}[F(a) + \{F(b)-F(a)\}U] \sim X.$$


## Fundamental theorem of simulation

Consider a $d$-variate random vector $\boldsymbol{X},$ independent of $U \sim \mathsf{unif}(0,1)$ and $c>0$.
If $(\boldsymbol{X}, U)$ is uniformly distributed on the set
\begin{align*}
\mathcal{A}_{f}=\{(\boldsymbol{x}, u): 0 \leq u \leq  c f(\boldsymbol{x})\},
\end{align*}
then $\boldsymbol{X}$ has density $f(\boldsymbol{x}).$ 
<!--
Conversely, if $\boldsymbol{X}$ has density $f(\boldsymbol{x})$ and $U\sim\mathsf{unif}(0,1)$ independently, then $[\boldsymbol{X}, cUf(\boldsymbol{X})]$ is uniformly distributed on $\mathcal{A}_f$
-->

- $f$ is the marginal density of $\boldsymbol{X}$ since $f(\boldsymbol{x}) = \int_0^{f(\boldsymbol{x})} \mathrm{d} u.$
- If we can simulate uniformly from $\mathcal{A}_{f},$ then, we can discard the auxiliary variable $u.$ See @Devroye:1986, Theorem 3.1.

## Fundamental theorem of simulation in picture


```{r}
#| eval: true
#| echo: false
#| label: fig-fundamental-theorem
#| fig-cap: "Illustration of the fundamental theorem of simulation. All points in blue below the density curve belong to $\\mathcal{A}_f.$"
a = 4
b = 1.25
# curve(dbeta(x, shape1 = a, shape2 = b))
mode <- (a-1)/(a+b-2)
n <- 300
height <- dbeta(mode, a, b) + 0.1
x <- runif(n)
u <- runif(n) * height
inside <- u < dbeta(x, a, b)
# abline(v = mode, lty = "dashed")
ggplot() +
  stat_function(fun = dbeta, args = c(shape1 = a, shape2 = b), xlim = c(0,1), n = 1001) +
  geom_point(data.frame(u = u, x = x, col = factor(inside)),
               mapping = aes(x = x, y = u, color = col), show.legend = FALSE) +
  MetBrewer::scale_color_met_d("Hiroshige") +
  scale_x_continuous(limits = c(0,1), expand = expansion(add = c(0.02,0.02)), breaks = seq(0, 1, by = 0.25), labels = c("0","0.25","0.5","0.75","1")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0,0.02))) +
  theme_classic()
```


## Simulation algorithms: accept-reject



- **Target**: sample from density $p(x)$ (hard to sample from)
- **Proposal**: find a density $q(x)$ with nested support, $\mathrm{supp}(p) \subseteq \mathrm{supp}(q)$, such that for all $x \in \mathrm{supp}(p)$, 
$$\frac{p(x)}{q(x)} \leq C, \quad C \geq 1.$$
*Uses the fundamental theorem of simulation by finding an envelope $Cq(x)$ that is over the density $p(x).$*

## Rejection sampling algorithm

1. Generate $X$ from proposal with density $q(x)$.
2. Compute the ratio $R \gets p(X)/ q(X)$.
3. If $CU \leq R$ for $U \sim \mathsf{unif}(0,1)$, return $X$, else go back to step 1.


## Accept-reject illustration


```{r}
#| eval: true
#| echo: false
#| label: fig-acceptreject
#| fig-cap: "Target density (full) and scaled proposal density (dashed): the vertical segment at $x=1$ shows the percentage of acceptance for a uniform slice under the scaled proposal, giving an acceptance ratio of 0.58."
ptarget <- function(x){
  0.3*TruncatedNormal::dtmvnorm(x = cbind(x), mu = 2, sigma = matrix(0.25),lb = 0) + 
    0.4*TruncatedNormal::dtmvnorm(x = cbind(x), mu = 0, sigma = matrix(0.5),lb = 0, ub = 3) + 0.3*TruncatedNormal::dtmvnorm(x = cbind(x), mu = 0, sigma = matrix(4),lb = 0, ub = 5)
}

Cst <- dgamma(x = 4, shape = 4, rate = 1/0.75)/ptarget(2)
qprop <- function(x){1.02*dgamma(x = x+2, shape = 4, rate = 1/0.75)/Cst}
# curve(ptarget, from = -0.2, to = 4, n = 1001, ylim = c(0, 2))
# curve(qprop, from = 0, to = 4, add = TRUE, lty = 2)
ggplot() +
  stat_function(fun = ptarget, n = 1001, xlim = c(0, 5)) +
  stat_function(fun = qprop, n = 1001, xlim = c(0, 5), linetype = "dashed") +
  geom_segment(data = data.frame(
    x0 = 1, x1 = 1, y0 = 0, y1 = qprop(1)),
               mapping = aes(
                 x = x0, y = y0, xend = x1, yend = y1), 
    linewidth = 2) +
  geom_segment(data = data.frame(
    x0 = 1, x1 = 1, y0 = 0, y1 = ptarget(1)),
               mapping = aes(
                 x = x0, y = y0, xend = x1, yend = y1),
    linewidth = 2, col = "grey") +
  labs(x = "x", 
       y = "", 
       subtitle = "scaled proposal and target densities") +
    scale_y_continuous(limits = c(0, 0.62), expand = c(0,0)) + 
    scale_x_continuous(expand = c(0,0)) + 
  theme_classic() 
```


## Remarks on rejection sampling

- Acceptance rate is $1/C$:
   - we need on average $C$ draws from $q$ to get one from $p$.
- $q$ must be more heavy-tailed than $p$
   - e.g., $q(x)$ Student-$t$ for $p(x)$ Gaussian.
- $q$ should be cheap and easy to sample from!

## Designing a good proposal density

Good choices must satisfy the following constraints: 

- pick a family $q(x)$ so that $$C = \mathrm{argmax}_x{p(x)}/{q(x)}$$ is as close to 1 as possible.
- you can use numerical optimization with target $$\log R(x) =\log p(x) - \log q(x)$$ to find the mode $x^\star$ and the upper bound $C = R(x^\star).$
- if $q(x; \boldsymbol{\theta})$ is a function of the parameter $\boldsymbol{\theta}$, search for $$\mathrm{argmin}_{\boldsymbol{\theta}} \mathrm{argmax}_x \{\log p(x) - \log q(x; \boldsymbol{\theta})\}$$

## Truncated Gaussian via accept-reject

Consider sampling $Y \sim \mathsf{Gauss}(\mu, \sigma^2)$, but truncated in the interval $(a, b)$. The target density is
\begin{align*}
p(x; \mu, \sigma, a, b) = \frac{1}{\sigma}\frac{\phi\left(\frac{x-\mu}{\sigma}\right)}{\Phi(\beta)-\Phi(\alpha)}.
\end{align*}
for $\alpha= (a-\mu)/\sigma$ and $\beta = (b-\mu)/\sigma$.
where $\phi(\cdot), \Phi(\cdot)$ are respectively the density and distribution function of the standard Gaussian distribution.

## Accept-reject (crude version)

1. Simulate $X \sim \mathsf{Gauss}(\mu, \sigma^2)$
2. reject any draw if $X < a$ or $X> b$. 

The acceptance rate is $C^{-1} = \{\Phi(\beta) - \Phi(\alpha)\}$

```{r}
#| eval: true
#| echo: true
# Standard Gaussian truncated on [0,1]
candidate <- rnorm(1e5)
trunc_samp <- candidate[candidate >= 0 & candidate <= 1]
# Acceptance rate
length(trunc_samp)/1e5
# Theoretical acceptance rate
pnorm(1)-pnorm(0)
```

## Accept-reject for truncated Gaussian  {.smaller}

Since the Gaussian is a location scale family, the inversion method gives
\begin{align*}
X \sim \mu + \sigma\Phi^{-1}\left[\Phi(\alpha) + \{\Phi(\beta)-\Phi(\alpha)\}U\right]
\end{align*}

We however need to evaluate $\Phi$ numerically (no closed-form expression).

The method fails for *rare event* simulation because the computer returns

- $\Phi(x) = 0$ for $x \leq -39$
- $\Phi(x)=1$ for $x \geq 8.3$,

implying that $a \leq 8.3$ for this approach to work [@LEcuyer.Botev:2017].

## Simulating tails of Gaussian variables {.smaller}

We consider simulation from a standard Gaussian truncated above $a>0$

Write the density of the truncated Gaussian as  [@Devroye:1986, p.381]$$f(x) = \frac{\exp(-x^2/2)}{\int_{a}^{\infty}\exp(-z^2/2)\mathrm{d} z}  =\frac{\exp(-x^2/2)}{c_1}.$$

Note that, for $x \geq a$, $x/a > 1$ and 
$$c_1f(x) \leq \frac{x}{a}\exp\left(-\frac{x^2}{2}\right)= a^{-1}\exp\left(-\frac{a^2}{2}\right)g(x);$$
where $g(x)=x\exp\{(a^2-x^2)/2\}\mathsf{1}_{x \ge a}$ is the density of a Rayleigh variable shifted by $a$.

::: {style="font-size: 50%;"}

The constant $C= \exp(-a^2/2)(c_1a)^{-1}$ approaches 1 quickly as $a \to \infty$ (asymptotically optimal).


:::

## Accept-reject: truncated Gaussian with Rayleigh  {.smaller}

The shifted Rayleigh has distribution function $$G(x) = 1-\exp\{(a^2-x^2)/2\}, x \geq a.$$ 




:::{.callout-important}

## Marsaglia algorithm
1. Generate a shifted Rayleigh  above $a$, $X \gets  \{a^2 - 2\log(U)\}^{1/2}$ for $U \sim \mathsf{unif}(0,1)$
2. Accept $X$ if $XV \leq a$, where $V \sim \mathsf{unif}(0,1)$.
:::

For sampling on $[a,b]$ with $a$ very large, propose from a Rayleigh truncated above at $b$ [@LEcuyer.Botev:2017].

```{r}
#| eval: true
#| echo: true
#| label: marsaglia-algo
a <- 8.3
niter <- 1000L
X <- sqrt(a^2 + 2*rexp(niter))
samp <- X[runif(niter)*X <= a]
```


::: {style="font-size: 80%;"}

Accept if $U \leq \frac{f(x)}{g(x)} \times \frac{1}{C} = \frac{\exp(-x^2/2)/c_1}{x\exp\{(a^2-x^2)/2\}} \times \exp(a^2/2)c_1a = \frac{a}{x}.$


:::

## Monte Carlo integration 

Calculate $\mathsf{E}_{X}\{g(X)\}$ of $X$ with density function $p$ on $\mathcal{X}.$ 

Monte Carlo integration proceeds by drawing $B$ independent samples $x_1, \ldots, x_B$ from $p$ on $\mathcal{X}$ and evaluating the empirical average of $g,$
\begin{align*}
\mathsf{E}\{g(X)\} = \int_{\mathcal{X}} g(x) p(x) \mathrm{d} x \approx \widehat{\mathsf{E}}\{g(X)\}=\frac{1}{B}\sum_{b=1}^B g(x_b).
\end{align*}


## Importance sampling

We can calculate the integral $\mathsf{E}\{g(X)\} = \int_{\mathcal{X}} g(x) p(x) \mathrm{d}x$ by considering instead draws from a density $q(\cdot)$ if $\mathcal{X} \subseteq \mathrm{supp}(q).$ Then,
\begin{align*}
\mathsf{E}\{g(X)\} = \int_{\mathcal{X}} g(x) \frac{p(x)}{q(x)} q(x) \mathrm{d} x.
\end{align*}


## Monte Carlo estimator for importance sampling

An alternative Monte Carlo estimator, which is biased but has lower variance, is obtained by drawing independent $x_1, \ldots, x_B$ from $q$ and taking instead the weighted average of
\begin{align*}
\widetilde{\mathsf{E}}\{g(X)\} =\frac{B^{-1} \sum_{b=1}^B w_b g(x_b) }{B^{-1}\sum_{b=1}^B w_b}.
\end{align*}
with weights $w_b = p(x_b)/q(x_b).$ The latter equal 1 on average, so one could omit the denominator without harm.


## Importance sampling for the variance of a beta distribution

Consider $X \sim \mathsf{beta}(\alpha, \alpha)$ for $\alpha > 1$, with expectation $\mathsf{E}(X)=0.5$ since the density is symmetric. 

The variance is $$\mathsf{Va}(X) = \mathsf{E}\{(X - 0.5)^2\}=\frac{1}{4 \cdot (2\alpha+1)}.$$ 


We consider an equiweighted (bimodal) mixture of $\mathsf{beta}(\alpha, 3\alpha)$ and $\mathsf{beta}(3\alpha, \alpha)$ as importance sampling density.



## Importance sampling for variance of a beta distribution

```{r}
#| eval: true
#| echo: false
#| warning: false
#| message: false
#| out-width: '95%'
#| fig-align: 'center'
B <- 2e4L; alpha <- 1.5; factor <- 3
# Mode at the mean 0.5
X0 <- rbeta(n = B, alpha, alpha)
px <- function(x){dbeta(x, alpha, alpha)}
# Importance sampling density - mixture of two betas (alpha, factor*alpha)
X1 <- ifelse(
  runif(B) < 0.5,
  rbeta(B, alpha, factor*alpha), 
  rbeta(B, factor*alpha, alpha))
qx <- function(x){
  0.5*dbeta(x, alpha, factor*alpha) + 0.5*dbeta(x, factor*alpha, alpha)}
# Function to integrate - gives variance of a symmetric beta distribution
g <- function(x){(x - 0.5)^2}
# Weights for importance sampling
w <- px(X1)/qx(X1)
# Compute quantities in a sequential manner
mc_est_seq <- cumsum(g(X0))/seq_len(B)
mc_var_seq <- sapply(1:B, function(i){var(g(X0[1:i]))/i})
# Importance sampling estimator
is_est_seq <- cumsum(g(X1)*w)/cumsum(w)
is_var_seq <- sapply(1:B, function(i){sum(w[1:i]^2*(g(X1[1:i]) - is_est_seq[i])^2)/ sum(w[1:i])^2})
mc_est <- mean(g(X0))
mc_var <- var(g(X0))/B
# Importance sampling weighted mean and variance
is_est <- weighted.mean(g(X1), w = w) # equivalent to mean(g(X1)*w)/mean(w)
is_var <- sum(w^2*(g(X1) - is_est)^2)/ (sum(w)^2)
# True value for the beta variance
th_est <- 1/(4*(2*alpha+1))
# Function to integrate
fn <- function(x){(x - 0.5)^2 * dbeta(x, alpha, alpha)}
palette <- MetBrewer::met.brewer(name = "Hiroshige",n = 2)
df1 <- data.frame(
  nsim = 1:B,
  ymc = mc_est_seq,
  yis = is_est_seq,
  interv1 = qnorm(0.975)*sqrt(mc_var_seq),
  interv2 = qnorm(0.975)*sqrt(is_var_seq))
g1 <- ggplot(data = df1[-(1:100),]) +
  geom_ribbon(mapping = aes(x = nsim,
                            y = ymc,
                            ymin = ymc - interv1,
                            ymax = ymc + interv1),
              alpha = 0.4,
              color = palette[1],
              fill = palette[1]) +
  geom_ribbon(mapping = aes(x = nsim,
                            y = yis,
                            ymin = yis - interv2,
                            ymax = yis + interv2),
              alpha = 0.4,
              color = palette[2],
              fill = palette[2]) +
  geom_hline(yintercept = th_est) +
  geom_line(aes(x = nsim, y = ymc), color = palette[1]) +
  geom_line(aes(x = nsim, y = yis), color = palette[2]) +
  scale_x_continuous(trans = "sqrt",
                     labels = scales::unit_format(unit = "K", scale = 1e-3)) +
  scale_y_continuous(limits = c(0.04, 0.08), expand = expansion()) +
  labs(x = "number of simulations",
       y = "",
       subtitle = "Variance estimation") +
  theme_classic()
g2 <- ggplot() +
  stat_function(fun = function(x){fn(x)*px(x)}, xlim = c(0,1)) +
  scale_x_continuous(limits = c(0,1),
                     breaks = seq(0, 1, by = 0.25),
                     labels = c("0","0.25","0.5","0.75","1")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0, 0.01))) +
  labs(x = "x", y = "", subtitle = expression(f(x)==(x-0.5)^2*p(x)))+
  theme_classic()
g3 <- ggplot() +
  stat_function(fun = px, xlim = c(0,1), color = palette[1]) +
  stat_function(fun = qx, linetype = "dashed", color = palette[2], xlim = c(0,1)) +
  scale_x_continuous(limits = c(0,1),
                     breaks = seq(0, 1, by = 0.25),
                     labels = c("0","0.25","0.5","0.75","1")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0, 0.1))) +
  labs(x = "x", y = "", subtitle = "density of Monte Carlo generator")+
  theme_classic()
g4 <- ggplot() +
  stat_function(fun = function(x){px(x)/qx(x)}, xlim = c(0,1)) +
  geom_segment(data = data.frame(x = X1[1:50], y = w[1:50]), mapping = aes(x = x, yend = y, y = 0)) +
  scale_x_continuous(limits = c(0,1),
                     breaks = seq(0, 1, by = 0.25),
                     labels = c("0","0.25","0.5","0.75","1")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(add = c(0,0.1))) +
  labs(y = "", x = "x", subtitle = "weights")+
  theme_classic()
(g2 + g3) / (g4 + g1) +
  theme_classic()
# par(mfrow = c(1,3), bty = "l")
# curve(fn, from = 0, to = 1, ylab = expression(f[n](x)), ylim = c(0, .125), yaxs ="i")
# curve(px, from = 0, to = 1, ylim = c(0, 1.4), yaxs = "i", xlim = c(0,1), xaxs = "i")
# curve(qx, from = 0, to = 1, add = TRUE, lty = 2, col = "gray")
## Weight curve
# curve(expr = px(x)/qx(x), from = 0, to = 1, n = 1001, xlim = c(0,1), xaxs = "i", ylab = "weights")
# points(X1[1:50], w[1:50], type = "h")

```

## Code for importance sampling of beta variance

```{r}
#| eval: true
#| echo: true
B <- 2e4L; alpha <- 1.5; factor <- 3
# Mode at the mean 0.5
X0 <- rbeta(n = B, alpha, alpha)
px <- function(x){dbeta(x, alpha, alpha)}
# Importance sampling density - mixture of two betas (alpha, factor*alpha)
X1 <- ifelse(
  runif(B) < 0.5,
  rbeta(B, alpha, factor*alpha), 
  rbeta(B, factor*alpha, alpha))
qx <- function(x){
  0.5*dbeta(x, alpha, factor*alpha) + 0.5*dbeta(x, factor*alpha, alpha)}
# Function to integrate - gives variance of a symmetric beta distribution
g <- function(x){(x - 0.5)^2}
# Weights for importance sampling
w <- px(X1)/qx(X1)
```

## Code for importance sampling of beta variance (continued)

```{r}

# Monte Carlo integration
mc_est <- mean(g(X0))
mc_var <- var(g(X0))/B
# Importance sampling weighted mean and variance
is_est <- weighted.mean(g(X1), w = w) # equivalent to mean(g(X1)*w)/mean(w)
is_var <- sum(w^2*(g(X1) - is_est)^2)/ (sum(w)^2)
# True value for the beta variance
th_est <- 1/(4*(2*alpha+1))
# Point estimates and differences
round(c(true = th_est, "monte carlo" = mc_est, "importance sampling" = is_est),4)
# Ratio of std. errors for means
mc_var/is_var # value > 1 means that IS is more efficient
```


## Markov chains

Plain ordinary Monte Carlo is great, but few algorithms are generic enough to be useful in complex high-dimensional problems.

We will instead typically build Markov chains that target an invariant stationary distribution. 


## Caveats?

Markov chain Monte Carlo methods generate **correlated** draws. 

**Questions**: 

1. can we use them as ordinary independent samples? 
2. what is the price to pay?

We need to do a little theoretical detour to answer these questions.

## Stationarity and Markov property


A stochastic process is

- **(strongly) stationary** if the distribution of $\{X_1, \ldots, X_t\}$ is the same as that of $\{X_{n+1}, \ldots X_{t+n}\}$ for any value of $n$ and given $t$.
- **weakly stationary** if $\mathsf{E}(X_t) = \mu$ for all $t$, and $\mathsf{Cov}(X_t, X_{t+h}) = \gamma_h$ does not depend on $t$.
- **Markov** if it satisfies the Markov property: given the current state of the chain, the future only depends on the current state and not on the past.

::: {style="font-size: 65%;"}
Strong stationarity implies weak stationarity.
:::


## Law of large number (ergodic theorem)

Let $\{Y_t\}$ is a weakly stationary sequence with mean $\mathsf{E}(Y_t)=\mu$ and $\gamma_h = \mathsf{Cov}(Y_t, Y_{t+h})$. Then, if the autocovariance series is convergent, meaning
$$\sum_{t=0}^\infty |\gamma_h| < \infty,$$
then $\{Y_t\}$ is ergodic for the mean and $\overline{Y} \stackrel{\mathrm{p}}{\to} \mu$.


Ergodicity means that two segments of a time series far enough apart act as independent.


## Ergodicity and transformations

The ergodic theorem is a law of large numbers for stochastic processes that allows for serial dependence between observations, provided the latter is not too large.

Any transformation $g(\cdot)$ of a stationary and ergodic process $\{Y_t\}$ retains the properties, so $\overline{g} = T^{-1} \sum_{t=1}^T g(Y_t) \to \mathsf{E}\{g(Y_t)\}$ as $T \to \infty.$





## Autoregressive process of order 1

Consider a first-order autoregressive process, or $\mathsf{AR}(1)$, 

$$Y_t = \mu + \phi(Y_{t-1} - \mu) + \varepsilon_t,$$ where 

- $\phi$ is the lag-one correlation, 
- $\mu$ the global mean
- $\varepsilon_t$ is an iid innovation with mean zero and variance $\sigma^2$

If $|\phi| < 1$, the process is stationary, otherwise the variance increases with $t$.

## Unconditional moments via tower law

If the process is weakly stationary, then $\mathsf{E}_{Y_{t}}(Y_t)=\mathsf{E}_{Y_{t-1}}(Y_{t-1})$
\begin{align*}
\mathsf{E}_{Y_{t}}(Y_t) &= \mathsf{E}_{Y_{t-1}}\left\{\mathsf{E}_{Y_{t} \mid Y_{t-1}}(Y_t)\right\}
\\&= \mu(1-\phi) + \phi\mathsf{E}_{Y_{t-1}}(Y_{t-1})
\end{align*}
and so the unconditional mean is $\mu$.
For the variance, we have
\begin{align*}
\mathsf{Va}_{Y_{t}}(Y_t) &= \mathsf{E}_{Y_{t-1}}\left\{\mathsf{Va}_{Y_{t} \mid Y_{t-1}}(Y_t)\right\} + \mathsf{Va}_{Y_{t-1}}\left\{\mathsf{E}_{Y_{t} \mid Y_{t-1}}(Y_t)\right\}\\
& = \sigma^2 + \mathsf{Va}_{Y_{t-1}}\left\{\mu + \phi(Y_{t-1} - \mu)\right\}
\\&= \sigma^2 + \phi^2 \mathsf{Va}_{Y_{t-1}}(Y_{t-1}).
\end{align*}
and the unconditional variance is $$\mathsf{Va}(Y_t) = \tau^2 = \sigma^2/(1-\phi^2).$$


## Autocovariance and Markov property

The covariance at lag $h$, in terms of innovations, gives
\begin{align*}
\gamma_h = \mathsf{Co}(Y_t, Y_{t-h}) = \mathsf{Va}(\phi Y_{t-1}, Y_{t-h}) + \mathsf{Va}(\varepsilon_t, Y_{t-h}) = \phi \gamma_{h-1}
\end{align*}
since $\varepsilon_t$ is independent of the past. We thus find $$\rho_h = \phi^h, \qquad \gamma_h = \phi^h\mathsf{Va}(Y_t).$$


The $\mathsf{AR}(1)$ process is first-order Markov since the conditional distribution $p(Y_t \mid Y_{t-1}, \ldots, Y_{t-p})$ equals $p(Y_t \mid Y_{t-1}).$

## Variance of sample average

Intuitively, a sample of correlated observations carries less information than an independent sample of draws.

The variance of the sample average is
\begin{align*}
\mathsf{Va}\left(\overline{Y}_T\right) &= \frac{1}{T^2}\sum_{t=1}^T \sum_{s=1}^T \mathsf{Co}(Y_t, Y_s)
\\&= \frac{1}{T^2}\sum_{t=1}^T \mathsf{Va}(Y_t) + \frac{2}{T^2} \sum_{t=1}^{T-1}\sum_{s = t+1}^T \mathsf{Co}(Y_t, Y_s).
\end{align*}
Under independence and assuming stationarity, we get $\mathsf{Va}\left(\overline{Y}_T\right)=\sigma^2/T.$


## Variance of sample average, redux

If the second moments are finite, the scaled limiting variance of the sample mean simplifies to
\begin{align*}
\lim_{T \to \infty} T\mathsf{Va}\left(\overline{Y}_T\right) = \tau^2 \left\{1+2\sum_{t=1}^\infty \rho_t\right\}.
\end{align*}
which is a function of 

- the unconditional variance $\tau^2$
- the lag-$h$ autocorrelation, $\mathsf{Cor}(Y_{t}, Y_{t+h})=\rho_h.$


## Correlogram

For a weakly stationary process, it is instructive to look at the autocorrelation $\rho_h=\mathsf{Cor}(Y_t, Y_{t+h}).$ 

```{r}
#| eval: true
#| echo: false
#| fig-height: 4
#| fig-width: 10
#| label: fig-correlograms
#| fig-cap: "Correlogram of two Markov chains. These plots, often called acf or autocorrelation functions, show the lag-h sample autocorrelation against lag number."
chains <- cbind("chain 1" = arima.sim(model = list(ar = c(0.5,-0.1), 
                                       ma = c(0.4)), n = 10000),
                "chain 2" = arima.sim(model = list(ar = 0.99), n = 10000))
bayesplot::mcmc_acf(x = coda::as.mcmc(chains)) + theme_classic()
```



## Effective sample size

The effective sample size is, loosely speaking, the equivalent number of observations if the $B$ marginal posterior draws where independent and more formally
$$
\mathsf{ESS} = \frac{B}{\sum_{t=-\infty}^\infty \rho_t}= \frac{B}{\left\{1+2\sum_{t=1}^\infty \rho_t\right\}} 
$$ 
where $\rho_t$ is the lag $t$ correlation, since $\rho_{t} = \rho_{-t}$ and $\rho_0=1.$


## Effective sample size for AR(1)

The lag-$h$ correlation of the stationary autoregressive process of order 1 is $\phi^h$, so
$$1+2\sum_{h=1}^\infty \rho_h = 1 + 2 \left(\frac{1}{1-\phi}-1\right) = \frac{1+\phi}{1-\phi}.$$

whereas the unconditional variance of the $\mathsf{AR}(1)$ process is 

$$\tau^2 = \frac{\sigma^2}{1-\phi^2}.$$

## Inefficiency curve for AR(1) {.smaller}

```{r}
#| label: fig-ar1-variance
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Left: scaled asymptotic variance of the sample mean for AR(1) (full line) and independent observations with unit marginal variance (dashed). Right: variance ratio for positive correlations for selected range."
#| message: false
#| warning: false
#| echo: false
#| eval: true
var2 <- function(rho){(1+rho)/(1-rho)}
var1 <- function(rho){1}
var_ratio <- function(rho){var2(rho)/var1(rho)}
g1 <- ggplot() +
  stat_function(fun = var1, xlim = c(-0.9,0.9), n = 1001L, linetype = "dashed") +
  stat_function(fun = var2, xlim = c(-0.9,0.9), n = 1001L) +
  labs(y = "", x = expression(phi)) +
  theme_classic()
g2 <- ggplot() +
  stat_function(fun = var_ratio,
                xlim = c(0.5,0.9), 
                n = 1001L) +
  labs(y = "") + 
  theme_classic()  
g1 + g2
```


To get the same precision for the mean of $\mathsf{AR}(1)$ process with $\phi \approx 0.75$ than with i.i.d. data, we would need 7 times as many observations.

## Morale of the story 

The price to pay for having correlated samples is 

:::{style="font-size: 2em; color: #ff585d; text-align: center"}

**inefficiency**

:::

The higher the autocorrelation, the larger the variability of our estimators.


## Standard error of mean for first-order autoregressive process

```{r}
#| eval: true
#| echo: true
phi <- 0.8 # lag-one autocorrelation
sigma <- 2 # std. deviation of white noise
B <- 1e4L # chain length
x <- arima.sim( # generate AR(1)
  model = list(ar = phi),
  n = B,
  rand.gen = rnorm,
  sd = sigma
)
# Effective sample size / B
ess_factor_ar1 <- 1 / (-1 + 2 / (1 - phi))
# Unconditional variance tau_sq, obtained from var(x)
uvar_ar1 <- sigma^2 / (1 - phi^2)
# Theoretical std. error of sample mean
se_mean_ar1 <- sqrt(uvar_ar1 / B / ess_factor_ar1)
```

## Standard errors estimation via nonparametric spectral analysis

Suppose that $X_1, \ldots, X_B$ is a mean-stationary sequence.

@Heidelberger.Welch:1981 recall that the variance of the sample mean is $S(0) / B,$ where $S(\cdot)$ is the spectrum evaluated at zero and
$$
S(0) = \sum_{h=-\infty}^\infty \gamma_h.
$$
They propose to approximate $S(0)$ by

- calculating the raw periodogram (unsmoothed)
- fitting a gamma GLM to the log periodogram with a polynomial trend.

## Standard errors estimation via parametric spectral analysis

Any continuous spectral density function can be approximated by an $\mathsf{AR}(p)$ process (for $p$ potentially infinitely large).

For an autoregressive process of order $p$, the spectral density of the $\mathsf{AR}(p)$ is [@Percival.Walden:2020, eq. 446b]
$$
S(0) = \frac{\sigma^2}{\left(1 - \sum_{j=1}^p \phi_j\right)^2}.
$$
This suggests

- Fitting an $\mathsf{AR}(p)$ model (selecting $p$ via $\mathsf{AIC}$) using Yule--Walker or any other method.
- Calculating the spectral density at zero with the plug-ins $\widehat{\phi}_1, \ldots, \widehat{\phi}_p$ and $\widehat{\sigma}^2.$

## Spectral estimation in **R**

```{r}
# Heidelberg and Welch (1981)
coda::spectrum0(x)
# Code from coda::effectiveSize
ar_mod <- ar.yw(x)
suvar <- ar_mod$var.pred / (1 - sum(ar_mod$ar))^2
se3 <- sqrt(suvar / B) # variance of sample mean of AR(p)
ess3 <- var(x) / suvar * B
```

## Standard errors for posterior means using batch means

@Geyer:2011 recommends to segment the time series into batches

1. Break the chain of length $B$ (after burn in) in $K$ blocks of size $\approx K/B.$
2. Compute the sample mean of each segment.
3. Compute the standard deviation of the segments mean.
4. Rescale by $K^{-1/2}$ to get standard error of the global mean.

## Illustration of batch means



```{r}
#| label: fig-mcmc-batchmean
#| eval: true
#| echo: false
#| fig-cap: "Calculation of the standard error of the posterior mean using the batch method."
set.seed(123450)
chain <- as.numeric(arima.sim(model = list(ar = c(0.9)), n = 2000)) + 0.4
batchMean <- function(x, batchSize = 100){
  niter <- length(x)
    nbatch <- niter%/%batchSize
    niter <- nbatch * batchSize
    ibatch <- rep(1:nbatch, each = batchSize)[1:niter]
    batchMeans <- t(sapply(split(data.frame(x[1:niter]), ibatch),
        function(batch) apply(batch, 2, mean)))
    return(as.numeric(batchMeans))
}
ggplot() +
  geom_line(data = data.frame(t = 1:2000,
                         chain = chain),
       mapping = aes(x = t,
                     y = chain),
       alpha = 0.1) +
  geom_segment(data = data.frame(x0 = seq(1, 2000, by = 100),
                                 x1 = seq(100, 2000, by = 100),
                                 y = batchMean(chain, batchSize = 100)),
               mapping = aes(x = x0, y = y, yend = y, xend = x1),
               linetype = "dashed") +
  geom_hline(yintercept = mean(chain)) +
  labs(x = "iteration number", y = "") +
  theme_classic()
```



## Convergence of Markov chains

If a Markov chain is irreducible and acyclic, it has a unique stationary distribution.

- **irreducibility**: means that the chain can move from anywhere to anywhere, so it doesn't get stuck in part of the space forever.
- **acyclic**: cyclical chains loop around and visit periodically a state


The ergodic theorem holds even if the chain is cyclical.


## Examples of cyclical or reducible chains

Consider discrete Markov chains over the integers $1, 2, 3$ with transition matrices

$$
P_1 = \begin{pmatrix}
0.5 & 0.3 & 0.2 \\
0 & 0.4 & 0.6 \\
0 & 0.5 & 0.5
\end{pmatrix},
\quad
P_2 = \begin{pmatrix}
0 & 0 & 1 \\
1 & 0 & 0 \\
0 & 1 & 0
\end{pmatrix}.
$$
Chain 1 is reducible to $\{2, 3\}$, chain 2 is cyclical.



## Convergence and stationary distribution

Consider a transition $P$ on $1, \ldots, 5$ defined as $$
P = \begin{pmatrix}
\frac{2}{3} & \frac{1}{3} &  0 & 0 & 0 \\
\frac{1}{6} & \frac{2}{3} & \frac{1}{6} & 0 & 0 \\
0 & \frac{1}{6} & \frac{2}{3} & \frac{1}{6} & 0 \\
0 & 0 & \frac{1}{6} & \frac{2}{3} & \frac{1}{6} \\
0 & 0 & 0 &  \frac{1}{3}  & \frac{2}{3} \\
\end{pmatrix}
$$ The stationary distribution is the value of the row vector $\boldsymbol{p},$ such that $\boldsymbol{p} = \boldsymbol{p}\mathbf{P}$ for transition matrix $\mathbf{P}$: we get $\boldsymbol{p}=(1,2,2,2,1)/8.$


## Convergence of Markov chains


```{r}
#| label: fig-discrete-markov-chain
#| eval: true
#| echo: false
#| cache: true
#| fig-cap: "Discrete Markov chain on integers from 1 to 5, with traceplot of 1000 first iterations (left) and running mean plots of sample proportion of each state visited (right)."
set.seed(1234)
P3 <- rbind(c(4,2,0,0,0),
            c(1,4,1,0,0),
            c(0,1,4,1,0),
            c(0,0,1,4,1),
            c(0,0,0,2,4)) / 6
chain <- integer(1e5L)
state <- sample.int(n = 5, size = 1)
for(i in seq_along(chain)){
  state <- sample.int(n = 5, size = 1, prob = P3[state,])
  chain[i] <- state
}
library(ggplot2)
library(patchwork)
g1 <- ggplot(data = data.frame(iteration = 1:1000,
                               state = chain[1:1000]),
             mapping = aes(x = iteration, y = state)) +
  geom_line() +
  theme_classic() +
  labs(y = "", subtitle = "state")
iter <- seq(from = 100, to = length(chain), by = 100)
sub <- sapply(iter,
              function(n){
                as.numeric(table(chain[seq_len(n)])/n)
                }) 
g2 <- ggplot(
  data = data.frame(
    x = rep(iter, each =5),
    y = c(sub),
    state = factor(rep(1:5, length.out = length(sub)))),
  mapping = aes(x = x, group = state, color = state, y = y)) +
  geom_line() +
  scale_x_continuous(trans = "sqrt", labels = scales::unit_format(unit = "K", scale = 1e-3)) + 
  labs(y = "",
       subtitle = "probability of state", 
       x = "iteration") +
  MetBrewer:::scale_color_met_d("Hiroshige") +
  theme_classic() +
  theme(legend.position = "bottom")
g1 + g2
```

## Summary

- Monte Carlo methods often rely on easy-to-simulate distributions $q(x)$ to draw from a target $p(x).$
- Rejection sampling: find 

## References

