---
title: "Bayesian modelling"
author: "Léo Belzile"
subtitle: "Markov chain Monte Carlo"
date: today
date-format: "[Last compiled] dddd MMM D, YYYY"
eval: true
echo: true
cache: true
bibliography: MATH80601A.bib
format:
  revealjs:
    slide-number: true
    html-math-method: mathjax
    preview-links: auto
    theme: [simple, hecmontreal.scss]
    title-slide-attributes:
      data-background-color: "#ff585d"
    logo: "fig/logo_hec_montreal_bleu_web.png"
---


```{r}
#| include: false
#| eval: true
#| echo: false
hecbleu <- c("#002855")
fcols <- c(gris = "#888b8d",
           bleu = "#0072ce",
           aqua = "#00aec7",
           vert = "#26d07c",
           rouge = "#ff585d",
           rose = "#eb6fbd",
           jaune = "#f3d03e")
pcols <- c(gris = "#d9d9d6",
           bleu = "#92c1e9",
           agua = "#88dbdf",
           vert = "#8fe2b0",
           rouge = "#ffb1bb",
           rose = "#eab8e4",
           jaune = "#f2f0a1")
library(ggplot2)
theme_set(theme_classic())
library(patchwork)
knitr::opts_chunk$set(fig.retina = 3, collapse = TRUE)
options(digits = 3, width = 75)
```

## Markov chain Monte Carlo

We consider simulating from a distribution with associated density function proportional to $p(\cdot)$

- using an algorithm that generates a Markov chain ,
- without requiring knowledge of the normalizing factor.


We use a conditional density $q(\boldsymbol{\theta} \mid \boldsymbol{\theta}^{\text{cur}})$ to generate proposals from the current value $\boldsymbol{\theta}^{\text{cur}}.$



## Metropolis--Hastings algorithm 

Starting from an initial value $\boldsymbol{\theta}_0$: for $t=1, \ldots, T$

1.  draw a proposal value $\boldsymbol{\theta}_t^{\star} \sim q(\boldsymbol{\theta} \mid \boldsymbol{\theta}_{t-1})$.
2.  Compute the acceptance ratio $$
    R = \frac{p(\boldsymbol{\theta}_t^{\star})}{p(\boldsymbol{\theta}_{t-1})}\frac{q(\boldsymbol{\theta}_{t-1} \mid \boldsymbol{\theta}_t^{\star} )}{q(\boldsymbol{\theta}_t^{\star} \mid \boldsymbol{\theta}_{t-1})}
    $$
3.  With probability $\alpha=\min\{R, 1\}$, accept the proposal and set $\boldsymbol{\theta}_t \gets \boldsymbol{\theta}_t^{\star}$, otherwise set the value to the previous state, $\boldsymbol{\theta}_t \gets \boldsymbol{\theta}_{t-1}$.

## Theory

The Metropolis--Hastings algorithm satisfies a technical condition (detailed balance) that ensures that the Markov chain generated is reversible. 

- Thus, any draw from the posterior will generate a new realization from the posterior. 
- Provided the starting value has non-zero probability under the posterior, the chain will converge to the stationarity distribution (albeit perhaps slowly).


## Interpretation



- If $R>1$, the proposal has higher density and we always accept the move. 
- If we reject the move, the Markov chain stays at the current value, which induces autocorrelation.
- Since the acceptance probability depends only on the density through ratios, normalizing factors of $p$ and $q$ cancel out.


## Symmetric proposals and random walk

If the proposal is symmetric, the ratio of proposal densities is $$q(\boldsymbol{\theta}_{t-1} \mid \boldsymbol{\theta}_t^{\star} ) / q(\boldsymbol{\theta}_t^{\star} \mid \boldsymbol{\theta}_{t-1}) = 1.$$

Common examples include random walk proposals
$$\boldsymbol{\theta}_t^{\star} \gets \boldsymbol{\theta}_{t-1} + \tau Z,$$ where $Z \sim \mathsf{Gauss}(0,1).$

## Independent proposals

- If we pick instead a global proposal, we must ensure that $q$ samples in far regions (recall rejection sampling), otherwise ...
- Good proposals include heavy tailed distribution such as Student-$t$ with small degrees of freedom, 

- centered at the maximum a posteriori $\widehat{\boldsymbol{\theta}}$ and 
- with scale matrix proportional to $-\mathbf{H}^{-1}(\boldsymbol{\theta}_t^{\star})$, where $\mathbf{H}(\cdot)$ is the Hessian of the (unnormalized) log posterior.


## Upworthy data example

We model the Poisson rates for headlines with questions or not. Our model is
\begin{align*}
Y_{i} &\sim \mathsf{Poisson}(n_i\lambda_i), \qquad (i=1,2)\\
\lambda_1 &= \exp(\beta + \kappa) \\
\lambda_2 &= \exp(\beta) \\
\beta & \sim \mathsf{Gauss}(\log 0.01, 1.5) \\
\kappa &\sim \mathsf{Gauss}(0, 1)
\end{align*}

## Implementation details: data and containers

In regression models, scale inputs if possible.

```{r}
#| eval: false
#| echo: true
# Load data
data(upworthy_question, package = "hecbayes")
# Compute sufficient statistics
data <- upworthy_question |>
  dplyr::group_by(question) |>
  dplyr::summarize(ntot = sum(impressions),
                   y = sum(clicks))
# Create containers for MCMC
niter <- 1e4L
chain <- matrix(0, nrow = niter, ncol = 2L)
colnames(chain) <- c("beta","kappa")
logpost_v <- numeric(niter)
```


## Implementation details: log posterior function

Perform all calculations on the log scale to avoid numerical overflow! 

```{r}
#| eval: false
#| echo: true
# Code log posterior as sum of log likelihood and log prior
loglik <- function(par, counts = data$y, offset = data$ntot, ...){
  lambda <- exp(c(par[1] + log(offset[1]), 
                  par[1] + par[2] + log(offset[2])))
 sum(dpois(x = counts, lambda = lambda, log = TRUE))
}
# Note common signature of function
logprior <- function(par, ...){
  dnorm(x = par[1], mean = log(0.01), sd = 1.5, log = TRUE) +
    dnorm(x = par[2], log = TRUE)
}
logpost <- function(par, ...){
  loglik(par, ...) + logprior(par, ...)
}
```

## Implementation details: proposals

Use good starting values for your Markov chains, such as maximum a posteriori.

```{r}
#| eval: false
#| echo: true
# Compute maximum a posteriori (MAP)
map <- optim(
  par = c(-4, 0.07),
  fn = logpost,
  control = list(fnscale = -1),
  offset = data$ntot,
  counts = data$y,
  hessian = TRUE)
# Use MAP as starting value
cur <- map$par
# Keep track of log posterior to reduce calculations
logpost_cur <- logpost(cur)
# Proposal covariance
cov_map <- -2*solve(map$hessian)
chol <- chol(cov_map)
```

## Implementation details: Metropolis--Hastings

Use seed for reproducibility, do not compute posterior twice, compute log of acceptance ratio. 

```{r}
#| eval: false
#| echo: true
set.seed(80601)
naccept <- 0L
for(i in seq_len(niter)){
  # Multivariate normal proposal - symmetric random walk
  prop <- c(rnorm(n = 2) %*% chol + cur)
  logpost_prop <- logpost(prop)
  logR <- logpost_prop - logpost_cur
  if(logR > -rexp(1)){# accept move
    cur <- prop
    logpost_cur <- logpost_v[i] <- logpost_prop
    naccept <- naccept + 1L
  }
  logpost_v[i] <- logpost_cur 
  chain[i,] <- cur
}
```

## Implementation details: analysis of output


```{r}
#| eval: false
#| echo: true
# Posterior summaries
summary(coda::as.mcmc(chain))
```

```
1. Empirical mean and standard deviation for each variable,
   plus standard error of the mean:

          Mean       SD  Naive SE Time-series SE
beta  -4.51268 0.001697 1.697e-05      6.176e-05
kappa  0.07075 0.002033 2.033e-05      9.741e-05

2. Quantiles for each variable:

          2.5%      25%      50%      75%    97.5%
beta  -4.51591 -4.51385 -4.51273 -4.51154 -4.50929
kappa  0.06673  0.06933  0.07077  0.07212  0.07463
```



## Metropolis--Hastings algorithm 

Starting from an initial value $\boldsymbol{\theta}_0$:

1.  draw a proposal value $\boldsymbol{\theta}_t^{\star} \sim q(\boldsymbol{\theta} \mid \boldsymbol{\theta}_{t-1})$.
2.  Compute the acceptance ratio $$
    R = \frac{p(\boldsymbol{\theta}_t^{\star})}{p(\boldsymbol{\theta}_{t-1})}\frac{q(\boldsymbol{\theta}_{t-1} \mid \boldsymbol{\theta}_t^{\star} )}{q(\boldsymbol{\theta}_t^{\star} \mid \boldsymbol{\theta}_{t-1})}
    $$
3.  With probability $\min\{R, 1\}$, accept the proposal and set $\boldsymbol{\theta}_t \gets \boldsymbol{\theta}_t^{\star}$, otherwise set the value to the previous state, $\boldsymbol{\theta}_t \gets \boldsymbol{\theta}_{t-1}$.

## Calculations


We compute the log of the acceptance ratio, $\ln R$, to avoid numerical overflow, with the log posterior difference
$$
\ln \left\{\frac{p(\boldsymbol{\theta}_t^{\star})}{p(\boldsymbol{\theta}_{t-1})}\right\} = \ell(\boldsymbol{\theta}_t^{\star}) + \ln p(\boldsymbol{\theta}_t^{\star}) - \ell(\boldsymbol{\theta}_{t-1}) - \ln p(\boldsymbol{\theta}_{t-1}) 
$$


Compare the value of $\ln R$ (if less than zero) to $\log(U)$, where $U \sim \mathsf{unif}(0,1)$. 


## What proposal?

The *independence* Metropolis--Hastings uses a global proposal $q$ which does not depend on the current state (typically centered at the MAP)

- This may be problematic with multimodal targets.

The Gaussian random walk takes $\boldsymbol{\theta}_t^{\star} =\boldsymbol{\theta}_{t-1}+ \sigma_\text{p}Z$, where $Z \sim \mathsf{Gauss}(0,1)$ and $\sigma_\text{p}$ is the proposal standard deviation. *Random walks* allow us to explore the space.


## Assessing convergence

How do we assess convergence of a MCMC algorithm?

- the algorithm implementation must be correct,
- the chain must have converged to the target posterior.
- the effective sample size must be sufficiently large for inference.


## Multiple chains



## Tracking the log posterior

One can store (separately) at each iteration the values of the 

- log likelihood
- log prior

These can be used for prior or likelihood sensitivity [@Kallioinen:2023]

## Typical set

We call the **typical set** the values that are taken mostly by the log posterior
The unnormalized


## Blank run

To check that the algorithm is correctly implemented:

- remove the log likelihood component
- run the algorithm with just the prior

The resulting draws should be drawn from the prior if the latter is proper [@Green:2001, p.55].



## Burn in

We are guaranteed to reach stationarity with Metropolis--Hastings, but it may take a large number of iterations...

One should discard initial draws during a **burn in** or warmup period if the chain has not reached stationarity. Ideally, use good starting value to reduce waste.

We can also use the warmup period to adapt the variance of the proposal.

## Goldilock principle and proposal variance

Mixing of the chain requires just the right variance (not too small nor too large).

```{r}
#| eval: true
#| echo: false
#| cache: true
#| label: fig-goldilock-trace
#| fig-cap: "Example of traceplot with proposal variance that is too small (top), adequate (middle) and too large (bottom)."
set.seed(80601)
niter <- 1000
fakesamp <- rnorm(n = 20, mean = 1, sd = 2)
fn <- function(par){ sum(dnorm(fakesamp, mean = par, sd = 2, log = TRUE))}
chains_goldi <- matrix(nrow = niter, ncol = 3)
cur <- rep(0,3)
for(i in seq_len(niter)){
  chains_goldi[i,] <-
    c(mgp::mh.fun(cur = cur[1], lb = -Inf, ub = Inf, prior.fun = identity, lik.fun = fn, pcov = matrix(0.01), cond = FALSE, transform = FALSE)$cur,
      mgp::mh.fun(cur = cur[2], lb = -Inf, ub = Inf, prior.fun = identity, lik.fun = fn, pcov = matrix(2), cond = FALSE, transform = FALSE)$cur,
      mgp::mh.fun(cur = cur[3], lb = -Inf, ub = Inf, prior.fun = identity, lik.fun = fn, pcov = matrix(25), cond = FALSE, transform = FALSE)$cur)
  cur <- chains_goldi[i,] 
}
accept <- apply(chains_goldi, 2, function(x){
  length(unique(x))/niter})
colnames(chains_goldi) <- c("small","medium","large")
# coda::traceplot(coda::as.mcmc(chains_goldi))
bayesplot::mcmc_trace(x = coda::as.mcmc(chains_goldi), 
           n_warmup = 0,facet_args = list(nrow = 3)) + theme_classic()

# chain1 <- mcmc::metrop(obj = fn, initial = 0, nbatch = 40, blen = 1000)
# chain2 <- mcmc::metrop(obj = fn, initial = 0, nbatch = 40, blen = 1000, scale = 0.1)
# chain3 <- mcmc::metrop(obj = fn, initial = 0, nbatch = 40, blen = 1000, scale = 0.4)

```

## Correlograms for Goldilock 


```{r}
#| eval: true
#| echo: false
#| fig-height: 4
#| fig-width: 10
#| label: fig-goldilock-correlogram
#| fig-cap: "Correlogram for the three Markov chains."
bayesplot::mcmc_acf(x = coda::as.mcmc(chains_goldi)) + theme_classic()
```


## Tuning Markov chain Monte Carlo

- Outside of starting values, the variance of the proposal has a huge impact on the asymptotic variance.
- We can adapt the variance during warmup by increasing/decreasing proposal variance (if acceptance rate is too large/small).
- We can check this via the acceptance rate (how many proposals are accepted).

## Optimal acceptance rates 

The following rules were derived for Gaussian targets under idealized situations.

- In 1D, rule of thumb is an acceptance rate of $0.44$ is optimal, and this ratio decreases to $0.234$ when $D \geq 2$ [@Sherlock:2013] for random walk Metropolis--Hastings.
- Proposals for $D$-variate update should have proposal variance of roughly $(2.38^2/d)\times \boldsymbol{\Sigma}$, where $\boldsymbol{\Sigma}$ is the posterior variance.
- For MALA (see later), we get $0.574$ rather than $0.234$

## Block update or one parameter at a time?

As with any accept-reject, proposals become inefficient when the dimension $D$ increase.

This is the **curse of dimensionality**.

Updating parameters in turn 

- increases acceptance rate (with clever proposals), 
- but also leads to more autocorrelation between parameters

## Solutions for strongly correlated coefficients

- Reparametrize the model to decorrelate variables (orthogonal parametrization).
- Block updates: draw correlated parameters together 
   - using the chain history to learn the correlation, if necessary



## Parameter transformation

Parameters may be bounded, e.g. $\theta_i \in [a,b]$.

- We can ignore this and simply discard proposals outside of the range, by setting the log posterior at $-\infty$ outside $[a,b]$
- We can do a transformation, e.g., $\log \theta_i$ if $\theta_i > 0$ and perform a random walk on the unconstrained space: don't forget Jacobians for $q(\cdot)$!
- Another alternative is to use truncated proposals (useful with more complex algorithms like MALA)


## Efficient proposals: MALA {.smaller}

The Metropolis-adjusted Langevin algorithm (MALA) uses a Gaussian random walk proposal $$\boldsymbol{\theta}^{\star}_t \sim \mathsf{Gauss}\{\mu(\boldsymbol{\theta}_{t-1}), \tau^2\mathbf{A}\},$$ with mean $$\mu(\boldsymbol{\theta}_{t-1})=\boldsymbol{\theta}_{t-1} + \mathbf{A}\eta \nabla \log p(\boldsymbol{\theta}_{t-1} \mid \boldsymbol{y}),$$ and variance $\tau^2\mathbf{A}$, for some mass matrix $\mathbf{A}$, tuning parameter $\tau>0$. 

The parameter $\eta < 1$ is a learning rate. This is akin to a Newton algorithm, so beware if you are far from the mode (where the gradient is typically large)!


## Higher order proposals

For a single parameter update $\theta$, a Taylor series expansion of the log posterior around the current value suggests using as proposal density a Gaussian approximation with [@Rue.Held:2005]

- mean $\mu_{t-1} = \theta_{t-1} - f'(\theta_{t-1})/f''(\theta_{t-1})$ and 
- precision $\tau^{-2} = -f''(\theta_{t-1})$

We need $f''(\theta_{t-1})$ to be negative!

This gives **local adaption** relative to MALA (global variance).

## Higher order and moves

For MALA and cie., we need to compute the density of the proposal also for the reverse move for the expansion starting from the proposal $\mu(\boldsymbol{\theta}_{t}^\star)$.

These methods are more efficient than random walk Metropolis--Hastings, but they require the gradient and the hessian (can be obtained analytically using autodiff, or numerically).

## Modelling individual headlines of **Upworthy** example

The number of conversions `nclick` is binomial with sample size $n_i=$`nimpression`.

Since $n_i$ is large, the sample average `nclick`/`nimpression` is approximately Gaussian, so write

\begin{align*}
Y_i &\sim \mathsf{Gauss}(\mu, \sigma^2/n_i)\\
\mu &\sim \mathsf{trunc. Gauss}(0.01, 0.1^2, 0, 1) \\
\sigma &\sim \mathsf{expo}(0.7)
\end{align*}

## MALA: data set-up

```{r}
#| eval: false
#| echo: true
data(upworthy_question, package = "hecbayes")
# Select data for a single question
qdata <- upworthy_question |>
  dplyr::filter(question == "yes") |>
  dplyr::mutate(y = clicks/impressions,
                no = impressions)
```

## MALA: define functions
```{r}
#| eval: false
#| echo: true
# Create functions with the same signature (...) for the algorithm
logpost <- function(par, data, ...){
  mu <- par[1]; sigma <- par[2]
  no <- data$no
  y <- data$y
  if(isTRUE(any(sigma <= 0, mu < 0, mu > 1))){
    return(-Inf)
  }
  dnorm(x = mu, mean = 0.01, sd = 0.1, log = TRUE) +
  dexp(sigma, rate = 0.7, log = TRUE) + 
  sum(dnorm(x = y, mean = mu, sd = sigma/sqrt(no), log = TRUE))
}
```

## MALA: compute gradient of log posterior

```{r}
#| eval: false
#| echo: true
logpost_grad <- function(par, data, ...){
   no <- data$no
  y <- data$y
  mu <- par[1]; sigma <- par[2]
  c(sum(no*(y-mu))/sigma^2 -(mu - 0.01)/0.01,
    -length(y)/sigma + sum(no*(y-mu)^2)/sigma^3 -0.7
  )
}
```


## MALA: compute maximum a posteriori

```{r}
#| eval: false
#| echo: true
# Starting values - MAP
map <- optim(
  par = c(mean(qdata$y), 0.5),
  fn = function(x){-logpost(x, data = qdata)},
  gr = function(x){-logpost_grad(x, data = qdata)},  
  hessian = TRUE,
  method = "BFGS")
# Check convergence 
logpost_grad(map$par, data = qdata)
```

## MALA: starting values and mass matrix

```{r}
#| eval: false
#| echo: true
# Set initial parameter values
curr <- map$par 
# Compute a mass matrix
Amat <- solve(map$hessian)
# Cholesky root - for random number generation
cholA <- chol(Amat)
```

## MALA: containers and setup

```{r}
#| eval: false
#| echo: true
# Create containers for MCMC
B <- 1e4L # number of iterations
warmup <- 1e3L # adaptation period
npar <- 2L
prop_sd <- rep(1, npar) # tuning parameter
chains <- matrix(nrow = B, ncol = npar)
damping <- 0.8
acceptance <- attempts <- 0 
colnames(chains) <- names(curr) <- c("mu","sigma")
# Proposal variance proportional to inverse hessian at MAP
prop_var <- diag(prop_sd) %*% Amat %*% diag(prop_sd)
```

## MALA: sample proposal with Newton step

```{r}
#| eval: false
#| echo: true
for(i in seq_len(B + warmup)){
  ind <- pmax(1, i - warmup)
  # Compute the proposal mean for the Newton step
  prop_mean <- c(curr + damping * 
     Amat %*% logpost_grad(curr, data = qdata))
  # prop <- prop_sd * c(rnorm(npar) %*% cholA) + prop_mean
  prop <- c(mvtnorm::rmvnorm(
    n = 1,
    mean = prop_mean, 
    sigma = prop_var))
#  [...]
```

## MALA: reverse step

```{r}
#| eval: false
#| echo: true
  # Compute the reverse step
  curr_mean <- c(prop + damping * 
     Amat %*% logpost_grad(prop, data = qdata))
  # log of ratio of bivariate Gaussian densities
  logmh <- mvtnorm::dmvnorm(
    x = curr, mean = prop_mean, 
    sigma = prop_var, 
    log = TRUE) - 
    mvtnorm::dmvnorm(
      x = prop, 
      mean = curr_mean, 
      sigma = prop_var, 
      log = TRUE) + 
  logpost(prop, data = qdata) - 
    logpost(curr, data = qdata)
```

## MALA: Metropolis--Hastings ratio

```{r}
#| eval: false
#| echo: true
  if(logmh > log(runif(1))){
    curr <- prop
    acceptance <- acceptance + 1L
  }
  attempts <- attempts + 1L
  # Save current value
  chains[ind,] <- curr
```

## MALA: adaptation

```{r}
#| eval: false
#| echo: true
  if(i %% 100 & i < warmup){
    # Check acceptance rate and increase/decrease variance
    out <- hecbayes::adaptive(
      attempts = attempts, # counter for number of attempts
      acceptance = acceptance, 
      sd.p = prop_sd, #current proposal standard deviation
      target = 0.574) # target acceptance rate
    prop_sd <- out$sd # overwrite current std.dev
    acceptance <- out$acc # if we change std. dev, this is set to zero
    attempts <- out$att # idem, otherwise unchanged
    prop_var <- diag(prop_sd) %*% Amat %*% diag(prop_sd)
  }
} # End of MCMC for loop
```


## Summary




## References

