---
title: "Bayesian modelling"
author: "Léo Belzile"
subtitle: "Bayesian regression"
date: today
date-format: "[Last compiled] dddd MMM D, YYYY"
eval: true
echo: true
cache: true
bibliography: MATH80601A.bib
format:
  revealjs:
    slide-number: true
    html-math-method: mathjax
    preview-links: auto
    theme: [simple, hecmontreal.scss]
    title-slide-attributes:
      data-background-color: "#ff585d"
    logo: "fig/logo_hec_montreal_bleu_web.png"
---


```{r}
#| include: false
#| eval: true
#| echo: false
hecbleu <- c("#002855")
fcols <- c(gris = "#888b8d",
           bleu = "#0072ce",
           aqua = "#00aec7",
           vert = "#26d07c",
           rouge = "#ff585d",
           rose = "#eb6fbd",
           jaune = "#f3d03e")
pcols <- c(gris = "#d9d9d6",
           bleu = "#92c1e9",
           agua = "#88dbdf",
           vert = "#8fe2b0",
           rouge = "#ffb1bb",
           rose = "#eab8e4",
           jaune = "#f2f0a1")
library(ggplot2)
theme_set(theme_classic())
library(patchwork)
knitr::opts_chunk$set(fig.retina = 3, collapse = TRUE)
options(digits = 3, width = 75)
```

## Bayesian regression

Same, same, but different...

- Generalized linear models, with distributional assumptions and link functions.
- We assign priors to $\boldsymbol{\beta}$ which
   - can provide shrinkage (regularization towards zero)
   - can enable variable selection (spike and slab)

## Model setup

Consider regression models with 

- model (or design) matrix $\mathbf{X} \in \mathbb{R}^{n \times p}$ 
- regression coefficients $\boldsymbol{\beta} = (\beta_1, \ldots, \beta_p)^\top \in \mathbb{R}^p$

## Ordinary linear regression model.

In the ordinary linear regression model, observations are independent and homoscedastic and 
\begin{align*}
\boldsymbol{Y} \mid \mathbf{X}, \boldsymbol{\beta}, \omega \sim \mathsf{Gauss}_n(\beta_0\mathbf{1}_n + \mathbf{X}\boldsymbol{\beta}, \omega^{-1}\mathbf{I}_n).
\end{align*}

The intercept $\beta_0$ receives special treatment, is always included. It is typically assigned an improper prior $p(\beta_0) \propto 1.$




## Decomposition of quadratic forms

For quadratic forms (in $\boldsymbol{x}$) with
\begin{align*}
& (\boldsymbol{x} - \boldsymbol{a})^\top \mathbf{A}(\boldsymbol{x} - \boldsymbol{a}) + (\boldsymbol{x} - \boldsymbol{b})^\top \mathbf{B}(\boldsymbol{x} - \boldsymbol{b}) \\
&\stackrel{\boldsymbol{x}}{\propto} (\boldsymbol{x} - \boldsymbol{c})^\top \mathbf{C}(\boldsymbol{x} - \boldsymbol{c})
\end{align*}
where $\mathbf{C} = \mathbf{A} + \mathbf{B}$ and $\boldsymbol{c}= \mathbf{C}^{-1}(\mathbf{A}\boldsymbol{a} + \mathbf{B}\boldsymbol{b})$.

This is useful to complete the square in Gaussian-Gaussian models.

## Bayesian Gaussian linear model

Consider Gaussian-gamma **conjugate** priors for the mean and precision parameters $\boldsymbol{\beta}$ and $\omega$, 
\begin{align*}
\boldsymbol{\beta} \mid \omega &\sim \mathsf{Gauss}\left\{\boldsymbol{\mu}_0, (
\omega\boldsymbol{\Omega}_0)^{-1}\right\} \\
\omega &\sim \mathsf{gamma}(\nu_0/2,\tau_0/2).
\end{align*}
Recall the sampling distribution of the ordinary least squares estimator is $$\widehat{\boldsymbol{\beta}} \sim \mathsf{Gauss}_p\{\boldsymbol{\beta}, (\omega\mathbf{X}^\top\mathbf{X})^{-1}\}.$$

## Conditional distributions

The conditional and marginal posterior distributions for the mean coefficients $\boldsymbol{\beta}$ and for the precision $\omega$ are
\begin{align*}
\boldsymbol{\beta} \mid \omega, \boldsymbol{y} &\sim \mathsf{Gauss}_p\left\{\boldsymbol{\mu}_n, (\omega\boldsymbol{\Omega}_n)^{-1}\right\}  \\
\omega \mid  \boldsymbol{y} &\sim \mathsf{gamma}\left\{(\nu_0 + n)/2,  \tau^2_n/2\right\}.
\end{align*}

If we integrate over the precision, we get instead
\begin{align*}
\boldsymbol{\beta} \mid  \boldsymbol{y} &\sim \mathsf{Student}_p(\boldsymbol{\mu}_n,  \tau_n/(\nu_0+n) \times \mathbf{\Omega}_n^{-1}, \nu_0 + n)
\end{align*}

## Posterior parameters


The precision is the sum of the precision of OLS estimator and prior precision.

The posterior mean is a weighted combination of the prior and OLS means, weighted by the scaled precision.
\begin{align*}
\boldsymbol{\Omega}_n &= \mathbf{X}^\top\mathbf{X} + \boldsymbol{\Omega}_0\\
\boldsymbol{\mu}_n &= \boldsymbol{\Omega}_n^{-1}(\mathbf{X}^\top\mathbf{X}\widehat{\boldsymbol{\beta}} + \boldsymbol{\Omega}_0\boldsymbol{\mu}_0) = \boldsymbol{\Omega}_n^{-1}(\mathbf{X}^\top\boldsymbol{y} + \boldsymbol{\Omega}_0\boldsymbol{\mu}_0)\\
\tau_n &= \tau_0 + (\boldsymbol{y} - \mathbf{X}\widehat{\boldsymbol{\beta}})^\top(\boldsymbol{y} - \mathbf{X}\widehat{\boldsymbol{\beta}}) + (\boldsymbol{\mu}_n - \widehat{\boldsymbol{\beta}})^\top \mathbf{X}^\top\mathbf{X}(\boldsymbol{\mu}_n - \widehat{\boldsymbol{\beta}}) \\& \quad + (\boldsymbol{\mu}_n-\boldsymbol{\mu}_0)^\top\boldsymbol{\Omega}_0(\boldsymbol{\mu}_n-\boldsymbol{\mu}_0)
\end{align*}


## Scale mixture of Gaussians

If $X \mid \sigma^2 \sim \mathsf{Gauss}(0, \sigma^2)$ and we assign a prior $p(\sigma^2)$

- if $\sigma^2 \sim \mathsf{inv. gamma}(\nu/2, \nu/2)$, then $X \sim \mathsf{Student}(0,1, \nu)$
- if $\sigma^2 \sim \mathsf{exp}(1/\lambda^2),$ then
$X \sim \mathsf{Laplace}(0, \lambda).$

## Sketch of proof 

1. write down the joint posterior as
\begin{align*}
 p(\boldsymbol{\beta}, \omega \mid \boldsymbol{y}) &\propto p(\boldsymbol{y} \mid \boldsymbol{\beta}, \omega) p(\omega)
\end{align*}
2. rewrite the first quadratic form in $\boldsymbol{y}-\mathbf{X}\boldsymbol{\beta}$ using the orthogonal decomposition
\begin{align*}
 (\boldsymbol{y}-\mathbf{X}\widehat{\boldsymbol{\beta}}) + (\mathbf{X}\widehat{\boldsymbol{\beta}} - \mathbf{X}\boldsymbol{\beta})
\end{align*}
3. pull terms  together and separate the conditional posterior $p(\boldsymbol{\beta} \mid \boldsymbol{y}, \omega)$ and $p(\omega \mid \boldsymbol{y})$

## Sketch of proof (continued)

4. use decomposition of quadratic forms with $\boldsymbol{a} = \widehat{\boldsymbol{\beta}}$, $\mathbf{A}=\mathbf{X}^\top\mathbf{X}$, $\boldsymbol{b} = \boldsymbol{\mu}_0$ and $\mathbf{B}=\boldsymbol{\Omega}_0$
5. the marginal of $\boldsymbol{\beta}$ is obtained by regrouping all terms that depend on $\omega$ and integrating over the latter, recognizing the integral as an unnormalized gamma density

## Cultural appropriation

Study 4 of @Lin.Kim.Uduehi.Keinan:2024 is a 3 by 2 by 2 three-way between-subject ANOVA focusing on cultural appropriation using a fictional scenario on publication of a soul food recipe cookbook from Chef Dax.

## Experimental variables

- ethnicity: chef is African-American or not
- action:  the way he obtained the recipes (by peeking without permission in kitchens, by asking permission or without mention (control)
- political ideology of respondant (liberal or conservative). 





## Posterior densities for marginal effects

```{r}
#| eval: true
#| echo: false
#| label: fig-contrasts
#| fig-cap: "Difference in appropriation rating for black vs non-black Chef Dax, average accross different levels of brand action."
ggplot(data = 
         data.frame(x = c(tc),
                    f = factor(rep(c("liberal","conservative"), 
                                   each = nrow(beta)))),
       mapping = aes(x = x, y = f, group = f)) +
  ggdist::stat_dist_halfeye(.width = 0.5) +
  labs(y = "",
       x = "difference in cultural appropriation score",
       subtitle = "Comparison when Chef Dax is black (vs not)") +
theme_classic()
```

## Chef Dax and cultural appropriation

The coefficients and standard errors from the linear regression are very nearly similar to the posterior mean and standard deviations for $\boldsymbol{\beta}$ from the marginal Student-$t,$ owing to the large sample size and uninformative priors.

On average, liberals perceive cultural appropriation more strongly (with nearly 2 points more), than conservatives (0.7 points on average).


```{r}
#| eval: true
#| echo: false
data(LKUK24_S4, package = "hecedsm")
options(contrasts = c("contr.sum", "contr.poly"))
model <- lm(appropriation ~ politideo * chefdax * brandaction,
          data = LKUK24_S4)
# Model matrix, response and dimensions
y <- LKUK24_S4$appropriation
X <- model.matrix(model)
n <- nrow(X)
p <- ncol(X)
# Priors
# We set zero for contrasts and a small value
# for the global mean (since response is on [1,7] Likert scale)
# with lower values indicating more cultural appropriation)
mu_0 <- c(2.5, rep(0, p-1))
Omega_0 <- diag(p)
# Prior precision of 0.25 (variance of 4)
nu_0 <- 0.25
tau_0 <- 1
Omega_n_inv <- solve(crossprod(X) + Omega_0)
mu_n <- Omega_n_inv %*% (crossprod(X, y) + crossprod(Omega_0, mu_0))
tau_n <- tau_0 + sum(resid(model)^2) + c(crossprod(X %*% (mu_n-coef(model)))) + sum((mu_n-mu_0)^2)
# Posterior draws from the model
omega <- rgamma(n = 1e3L, 
                shape = (nu_0 + n)/2, 
                tau_n^2/2)
beta <- matrix(rnorm(n = 1e3L*p), ncol = p) %*% 
  chol(Omega_n_inv)
beta <- sweep(beta, 1, sqrt(omega), "/")
beta <- sweep(beta, 2, mu_n, "+")
# Posterior quartiles for beta
beta_qu <- t(apply(beta, 2, quantile, probs = c(0.25,0.5,0.75)))
# Standard dev. for beta (from Student-t)
beta_se <- sqrt((nu_0 + n)/(nu_0 + n - 2) *diag(tau_n/(nu_0 + n) * Omega_n_inv))

beta <- TruncatedNormal::rtmvt(
  n = 1e3L,
  mu = mu_n, 
  sigma = tau_n/(nu_0 + n) * Omega_n_inv, 
  df = nu_0 + n)
dfg <- expand.grid(politideo =c("conservative", "liberal"),
            chefdax = c("not black", "black"),
            brandaction = c("peeking","permission", "control"))
mm <- model.matrix( ~ politideo * chefdax * brandaction,
              data = dfg)
# Subgroup means for each of the 12 categories
mu <- tcrossprod(beta, mm)
# Contrast weights, averaging over brandaction
w1 <- rep(c(1, 0, -1, 0), length.out = nrow(dfg))/3
w2 <- rep(c(0, 1, 0, -1), length.out = nrow(dfg))/3
# Posterior distribution of contrasts
tc <- mu %*% cbind(w1, w2)
```



## Modelling random effects

Gaussian mixed models in frequentist statistics are of the form
\begin{align*}
\boldsymbol{Y} \mid \mathcal{B}=\boldsymbol{b} &\sim \mathsf{Gauss}_n\left(\mathbf{X}\boldsymbol{\beta} + \mathbf{Z}\boldsymbol{b}, \sigma^2 \mathbf{I}_n\right)\\
\mathcal{B} &\sim \mathsf{Gauss}_q(\boldsymbol{0}_q, \boldsymbol{\Psi}).
\end{align*}

Bayesians also assign priors to $\boldsymbol{\beta}$ as well! but typically apriori independent with  $\mathsf{Va}_{\boldsymbol{\beta}}(\boldsymbol{\beta}) \propto \mathbf{I}_p.$

## Prior for covariance matrices

We need a prior for $p \times p$  symmetric positive definite  matrix random matrices!

We consider two cases: 

- Wishart (precision) / inverse Wishart (covariance), the conjugate prior for Gaussian
- onion peel prior on the correlation matrix


Wishart allows for conjugacy, but has unintuitive properties.

## Wishart distribution

We say $\mathbf{Q} \sim \mathsf{Wishart}_p(\nu, \mathbf{S})$ for $\nu>0$ degrees of freedom and scale $\mathbf{S}$ if it's density is proportional to
\begin{align*}
 f(\boldsymbol{Q}) \stackrel{\boldsymbol{Q}}{\propto} |\boldsymbol{Q}|^{(\nu-p-1)/2}\exp\left\{-\frac{\mathrm{tr}(\mathbf{S}^{-1}\boldsymbol{Q})}{2}\right\}, \quad \nu > p-1.
\end{align*}
where $|\cdot|$ denotes the determinant of the matrix and $\mathrm{tr}(\cdot)$ the trace operator. 

## Wishart distribution

The Wishart arises from considering $n \geq p$ independent and identically distributed mean zero Gaussian vectors $\boldsymbol{Y}_i \sim \mathsf{Gauss}_p(\boldsymbol{0}_p, \mathbf{S})$, where
\begin{align*}
 \sum_{i=1}^{\nu} \boldsymbol{Y}_i\boldsymbol{Y}_i^\top \sim \mathsf{Wishart}_p(\nu, \mathbf{S}).
\end{align*}

## Prior elicitation for Wishart

For prior elicitation, the mean of the Wishart is $\nu \mathbf{S}$

- $\nu$ is thus a prior sample size
- $\mathbf{S}$ is a scale matrix, often the identity matrix.

## Inverse Wishart

Consider a prior for the covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{Q}^{-1}$. Applying the change of variable formula, we get Jacobian $|\boldsymbol{\Sigma}|^{p+1}$, and so $\boldsymbol{\Sigma} \sim \mathsf{inv. Wishart}(\nu, \mathbf{S}^{-1}),$ with density \begin{align*}
 p(\boldsymbol{\Sigma}) \propto |\boldsymbol{\Sigma}|^{-(\nu+p+1)/2} \exp\left\{-\frac{1}{2} \mathrm{tr}\left(\boldsymbol{S}^{-1}\boldsymbol{\Sigma}^{-1}\right)\right\}
\end{align*}
with expectation $\mathbf{S}^{-1}(\nu-p-1)$ for $\nu > p+1.$


## Wishart as conjugate prior in Gaussian model

Consider $\boldsymbol{\mu} \sim \mathsf{Gauss}_p(\boldsymbol{\mu}_0, \boldsymbol{Q}^{-1})$ and $\boldsymbol{Q} \sim \mathsf{Wishart}_p(\nu, \mathbf{S})$ for $\nu \geq p$. Then,
\begin{align*}
 p(\boldsymbol{Q} \mid \boldsymbol{\mu}, \boldsymbol{\mu}_0) \propto &  |\boldsymbol{Q}|^{(\nu-p-1)/2}\exp\{-\mathrm{tr}(\mathbf{S}^{-1}\boldsymbol{Q})/2\} \\ &\times |\boldsymbol{Q}|^{1/2} \exp \left\{ -\frac{1}{2} (\boldsymbol{\mu}-\boldsymbol{\mu}_0)^\top\boldsymbol{Q}(\boldsymbol{\mu}-\boldsymbol{\mu}_0)\right\} 
\end{align*}
and thus 
$$ \boldsymbol{Q} \mid \boldsymbol{\mu}, \boldsymbol{\mu}_0 \sim \mathsf{Wishart}_p\{\nu + 1/2, \boldsymbol{S} + (\boldsymbol{\mu}-\boldsymbol{\mu}_0)(\boldsymbol{\mu}-\boldsymbol{\mu}_0)^\top\}.$$ 

## Conjugacy

Note that a $1 \times 1$ matrix is equal to it's trace, and the trace operator is invariant to cyclic of it's argument, meaning that
\begin{align*}
 (\boldsymbol{\mu}-\boldsymbol{\mu}_0)^\top\boldsymbol{Q}(\boldsymbol{\mu}-\boldsymbol{\mu}_0) = \mathrm{tr}\left\{ \boldsymbol{Q}(\boldsymbol{\mu}-\boldsymbol{\mu}_0)(\boldsymbol{\mu}-\boldsymbol{\mu}_0)^\top\right\}.
\end{align*}

## Properties of Wishart

The marginal precision for the Wishart variate are gamma distributed with the same degrees of freedom $\nu$. 

- there is a single parameter governing all marginal variances.

Moreover, the absolute value of the correlation and marginal variance parameters are negatively related [@Gelman:2013]. Large variance thus correspond to small correlations shrunk towards zero when the degrees of freedom increase.

## Wishart draws


```{r}
#| eval: true
#| echo: false
#| label: fig-draws-Wishart
#| fig-cap: "Prior draws from a bivariate inverse Wishart with identity scale matrix and $\\nu \\in \\{3, 20\\}$ degrees of freedom."
set.seed(80601)
prec <- rWishart(n = 1000, df = 3, Sigma = diag(2))
cor_a <- apply(prec, 3, function(x){cov2cor(solve(x))[1,2]})
var_a <- apply(prec, 3, function(x){solve(x)[1,1]})

prec <- rWishart(n = 1000, df = 20, Sigma = diag(2))
cor_b <- apply(prec, 3, function(x){cov2cor(solve(x))[1,2]})
var_b <- apply(prec, 3, function(x){solve(x)[1,1]})
draws <- data.frame(cor = c(cor_a, cor_b),
           logvar = log(c(var_a, var_b)),
           df = factor(paste0("df = ",rep(c(3, 20), each = 1000))))
ggplot(data = draws,
       mapping = aes(x = cor, y = logvar)) +
  geom_point() +
  facet_wrap(facets = ~df) +
  labs(x = expression("correlation"~rho),
       y = expression("log of marginal variance"~sigma^2[1])) +
theme_classic()
```

## Onion peel prior


A better alternative is to specify 

- different prior for each marginal scale $\sigma_j$ and 
- a prior on the correlation matrix $\mathbf{R}.$


For the latter, the onion peel or LKJ prior, named after the authors of @Lewandowski:2009, is $$p(\mathbf{R}) \propto |\mathbf{R}|^{\eta-1}, \eta>0$$ 

The case $\eta=1$ leads to uniform over the space of correlation matrices, and $\eta>1$ favours the identity matrix.



## Shrinkage and variable selection


With $p$ covariates, there are $2^p$ potential regression models. 

This is too many models to explore for $p$ large, and too many parameters relative to sample size $n$.

Two solutions:


- shrinkage priors: penalize small coefficients by shrinking towards zero via priors on $\boldsymbol{\beta}$
- Bayesian model averaging: assign prior to each model (different sets of covariates $\mathbf{X}$) and get a mixture of models.




## Spike-and-slab prior

The **discrete spike-and-slab prior** [@Mitchell.Beauchamp:1988] is a two-component mixture with

- the spike: a point mass $\delta_0$ or a vary narrow distribution centered at zero
- the slab, a diffuse distribution.

\begin{align*}
\beta_j \mid \gamma_j, \sigma^2 \sim \gamma_j \delta_0 + (1-\gamma_j)\mathsf{Gauss}(0, \sigma^2)
\end{align*}


## Prior for the spike-and-slab prior probability

Set independent and identically distributed conjugate prior for $\gamma_j \sim \mathsf{binom}(1, \omega),$ whence
\begin{align*}
p(\boldsymbol{\gamma} \mid \omega) = \prod_{j=1}^n \omega^{\gamma_j} (1-\omega)^{1-\gamma_j}
\end{align*}

Apriori, we set $\omega \in (0,1)$ as the proportion of the $p$ coefficients $\boldsymbol{\beta}$ that are zero (so $p(1-\omega)$ nonzero coefficients).



## Continuous spike-and-slab prior

@George.McCulloch:1993 replaced the spike by a Gaussian with near infinite precision around zero, with
\begin{align*}
 \beta_j \mid \gamma_j, \sigma_j^2,\phi^2_j \sim  \gamma_j \mathsf{Gauss}(0, \sigma_j^2\phi^2) + (1-\gamma_j) \mathsf{Gauss}(0, \sigma^2_j)
\end{align*}
where $\phi^2_j$ is very nearly zero, typically $\phi_j^2=0.001$. 

The construction allows for variable augmentation with mixture indicators and Gibbs sampling, although mixing tends to be poor.

## Horseshoe prior


The horseshoe prior of @Carvalho.Polson.Scott:2010 is a hierarchical prior of the form 
\begin{align*}
\beta_j \mid \sigma^2_j &\sim \mathsf{Gauss}(0, \sigma^2_j),\\\sigma^2_j \mid \lambda  &\sim \mathsf{Student}_{+}(0, \lambda, 1),\\ \lambda &\sim \mathsf{Student}_{+}(0, \omega, 1)
\end{align*}
where $\mathsf{Student}_{+}(0, a, 1)$ denotes a half-Cauchy distribution with scale $a>0,$ truncated on $\mathbb{R}_{+}.$


## Understanding shrinkage priors

The choice of $\sigma^2_j$ leads to an unconditional scale mixture of Gaussian for $\beta_j$. 

Better is to consider 
$$\kappa = 1 - 1/(1+\sigma^2) \in [0,1].$$
Penalization of near-zero components can be deduced from the density of $\kappa \to 0$, and similarly penalization of large signals by looking at the density when $\kappa \to 1.$




## Shrinkage weights
 
Weighting implied by Gaussian mixture density with Cauchy/Laplace/horsehoe.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-weights-shrinkage
#| fig-cap: "Density of penalization weights $\\kappa$ of spike (near zero) and slab (near one) for three shrinkage priors."
f1 <- function(x){exp(-0.5/x)/(x^2)}
f2 <- function(x){x^(-0.5)*(1-x)^(-1.5)*exp(-0.5*x/(1-x))}
x <- seq(0, 1, length.out = 1001)
f3 <- dbeta(x, 0.5, 0.5)

f1c <- integrate(f1, 0, 1)$value
f2c <- integrate(f2, 1e-8, 1)$value

prk_df <- data.frame( 
  density = c(f1(x)/f1c, f2(x)/f2c, f3), 
  x = rep(1-x, length.out = 3*length(x)), 
  prior = factor(rep(c("LASSO","Cauchy","horseshoe"), 
                     each = length(x))))
ggplot(data = prk_df,
       mapping = aes(x = x, y = density)) + 
 geom_line() +
 facet_wrap(~prior) +
scale_x_continuous(breaks = c(0, 0.5,1), labels = c("0","0.5","1"), expand = expansion()) + 
scale_y_continuous(limits = c(0, 5), expand = expansion()) +
theme_classic()
```

## Comparison of shrinkage priors

```{r}
#| eval: true
#| echo: false
#| cache: true
#| label: fig-shrinkage
#| fig-cap: "Marginal density for a regression coefficient $\\beta$ with horseshoe prior (full), Laplace (dashed) and a Student-$t$ (thick dotted)."
# Plot density of betas for various shrinkage


plam <- function(lambda, log = FALSE){
  dens <- 4*log(lambda) /(pi^2*(lambda^2-1))
  dens[lambda == 0] <- 0
  dens[lambda == 1] <- 2/(pi^2)
  if(isTRUE(log)){
    log(dens)
  } else{
   dens
  }
}

dhorseshoe <- function(pars, log = TRUE){
  logdens <- dnorm(pars[1], 0, sd = pars[2], log = TRUE) +
    plam(pars[2], log = TRUE)
  if(isTRUE(log)){
    logdens
  } else{
    exp(logdens)
  }
}

dhorseshoe_v <- function(lambda, beta, log = TRUE){
  if(length(lambda) > 1){
  stopifnot(length(beta) %in% c(1L, length(lambda)))
  }
  logdens <- dnorm(beta, 0, sd = lambda, log = TRUE) +
    plam(lambda, log = TRUE)
  if(isTRUE(log)){
    logdens
  } else{
    exp(logdens)
  }
}

B <- 1001L
beta_s <- seq(0.01, 6, length.out = B)
dbeta_hs1 <- dbeta_hs2 <- numeric(B)
for(i in seq_along(beta_s)){
  integ <- integrate(f = function(lam){
                 dhorseshoe_v(beta = beta_s[i], lambda = lam, log = FALSE)},
            lower = 0,
            upper = Inf)
  dbeta_hs1[i] <- integ$value
}
xs <- seq(-4,4, length.out = 101)
dl <- extraDistr::dlaplace(x = xs, mu = 0, sigma = 1)
par(mfrow = c(1,2), mar = c(5,5,1,1))
plot(x = c(-rev(beta_s), beta_s),
     y = c(rev(dbeta_hs1), dbeta_hs1),
     type = "l",
     ylab = "density",
     bty = "l",
     ylim = c(0,2),
     xlim = c(-4,4),
     yaxs = "i",
     xlab = expression(beta))
lines(xs, dl, lty = 2)
lines(xs, dt(xs, df = 1), lty = 3, lwd = 1.5)

xs2 <- seq(4, 6, length.out = 101)
plot(x = beta_s[beta_s>=4],
     y = dbeta_hs1[beta_s>=4],
     type = "l",
     ylab = "density",
     bty = "l",
     ylim = c(0,0.02),
     xlim = c(4,6),
     yaxs = "i",
     xlab = expression(beta))
lines(xs2, extraDistr::dlaplace(x = xs2, mu = 0, sigma = 1), lty = 2)
lines(xs2, dt(xs2, df = 1), lty = 3, lwd = 1.5)
```

## Comments about the horseshoe

While the horseshoe prior guarantees that large coefficients are not regularized, this feature of the shrinkage prior is harmful in certain instances, for example separation of variables for logistic regression.


Markov chain Monte Carlo simulations are hampered by these parameters whose posterior mean does not exist, leading to poor mixing. 

## Finnish horseshoe (aka regularized horseshoe)

@Piironen.Vehtari:2017 proposed instead
\begin{align*}
\beta_j \mid \lambda, \tau_j, c^2 &\sim \mathsf{Gauss}\left(0, \lambda\frac{c^2\tau_j^2}{c^2 + \tau^2_j\lambda^2}\right), \\
\tau_j &\sim \mathsf{Student}_{+}(0, 1, 1)\\
c^2 \mid s^2, \nu &\sim \mathsf{inv. gamma}(\nu/2, \nu s^2/2).
\end{align*}


## Shrinkage for Finnish horsshoe

When $\tau^2\lambda^2_j$ is much greater than $c^2$, this amounts to having a Student slab with $\nu$ degrees of freedom for large coefficients.

Taking a small value of $\nu$ allows for large, but not extreme components, and the authors use $s^2=2, \nu=4.$ 

## Hyperprior for Finnish horseshoe

The above specification does not specify the prior for the global scale $\lambda$, for which @Piironen.Vehtari:2017 recommend $$\lambda \sim \mathsf{Student}_{+}\left\{0, \frac{p_0}{(p-p_0)}\frac{\sigma}{n^{1/2}}, 1\right\},$$ where $p_0$ is a prior guess for the number of non-zero components out of $p,$ $n$ is the sample size and $\sigma$ is some level of the noise.


## Comparison of shrinkage priors

We revisit the `diabetes` data from the **R** package `lars`, which was used in @Park.Casella:2008 to illustrate the Bayesian LASSO. We consider three methods: 

- the default Gaussian prior, which gives a ridge penalty, 
- the Bayesian LASSO of @Park.Casella:2008
- the horseshoe prior.

Models are fitted using the `bayesreg` package.

## Density estimates of ordered coefficients 

```{r}
#| eval: true
#| echo: false
#| label: fig-shrinkage-dens-comparison
#| fig-cap: "Density estimates for regression coefficients with Gaussian (ridge), double exponential (Laplace) and horseshoe priors for the `diabetes` data."
data(diabetes, package = "lars")
set.seed(80601)
ridge <- bayesreg::bayesreg(y ~ scale(x), data = diabetes, prior = "ridge", thin = 1, n.samples = 2500L)
blasso <- bayesreg::bayesreg(y ~ scale(x), data = diabetes, prior = "lasso", thin = 1, n.samples = 2500L)
horseshoe <- bayesreg::bayesreg(y ~ scale(x), data = diabetes, prior = "hs", thin = 1, n.samples = 2500L)

df <- data.frame(rbind(t(ridge$beta), t(blasso$beta), t(horseshoe$beta)))
colnames(df) <- colnames(diabetes$x)
df$model <- factor(rep(c("ridge","lasso","horseshoe"), each = 2500L))
dfl <- tidyr::pivot_longer(df, cols = -model, names_to = "coef", values_to = "beta")
dfl <- dfl |>
  dplyr::mutate(coef = forcats::fct_reorder(coef,beta,  .fun='median'))
ggplot(data = dfl,
       mapping = aes(x = coef, y = beta, group = model, 
                      fill = model, color = model)) +   
geom_hline(yintercept = 0, col = "grey") + 
   ggdist::stat_dist_halfeye(position = "dodge", .width = c(0.5, 0.89)) +
MetBrewer::scale_color_met_d("Hiroshige") + 
MetBrewer::scale_fill_met_d("Hiroshige") + 
 labs(subtitle = expression(beta), y = "", x = "coefficient") +
  theme_classic() +
theme(legend.position = "bottom")
```

## Comments on penalization

- Ridge has the widest intervals of all methods, providing some shrinkage only for large values of $\beta$. 
- The horseshoe has typically narrower intervals, with more mass in a neighborhood of zero for smaller coefficients, and asymmetric intervals.
- The effective sample size fraction relative to the number of samples ranges from `r round(min(100*horseshoe$ess.frac), 0)`% to `r round(100*max(horseshoe$ess.frac), 0)`%, compared to `r round(100*min(blasso$ess.frac), 0)`% to `r round(100*max(blasso$ess.frac), 0)`% for the Bayesian LASSO and near-independent draws with the conjugate ridge.

## Bayesian model averaging

BMA refers to situation where we specify a mixture of models $M_1, \ldots,$, and we wish to recover the posterior weights of these.

This is useful for predictions (ensemble) methods to account for uncertainty in variable selection.

We consider design of MCMC for moving between models.

## Reversible jump MCMC

Reversible jump Markov chain Monte Carlo [@Green:1995] is an extension of the classical Metropolis--Hastings scheme that allows for arbitrary measures and through this varying dimensions.

Varying dimensions occurs not only with variable selection, but also changepoint analysis and mixture models with varying number of components.


Reversible jump requires **dimension-balancing**  and defining different types of moves for jumping between dimensions. 

## Dimension changes and jacobians

Dimensions changes are integrated in the Metropolis--Hastings step through a Jacobian term $J$: the probability of rejection $R$ for Metropolis becomes
\begin{align*}
    R = J\frac{p(\boldsymbol{\theta}_t^{\star})}{p(\boldsymbol{\theta}_{t-1})}\frac{q(\boldsymbol{\theta}_{t-1} \mid \boldsymbol{\theta}_t^{\star} )}{q(\boldsymbol{\theta}_t^{\star} \mid \boldsymbol{\theta}_{t-1})}
\end{align*}

In regression models, we will consider moves that adds or removes one parameter/regressor at a time.

## Setup

We consider models $M_1, \ldots, M_m$ with for simplicity $p(M_i)=1$ for all models that include an intercept.

We write $|M|$ for the cardinality of the set of non-zero coefficients $\boldsymbol{\beta}$ in model $M.$

Define $\mathbf{X}^{(m)}$ and $\boldsymbol{\beta}^{(m)}$ as the model matrix and the associated vector of non-zero coefficients associated with model $M_m$

## Setup for regression 
\begin{align*}
\boldsymbol{Y} \mid M_m, \boldsymbol{\beta}, \sim \mathsf{Gauss}(\mathbf{X}^{(m)}\boldsymbol{\beta}^{(m)}, \sigma^2 \mathbf{I}_n).
\end{align*}

We assign a Gaussian prior on $\boldsymbol{\beta}^{(m)} \mid M_m,$ is assigned a Gaussian prior, etc.

## Conditional Bayes factor


Write $\boldsymbol{\theta}$ for all parameters other than the response, model and vector of coefficients.

We can consider a joint update of the regression parameters $\boldsymbol{\beta}, M$ by sampling from their joint distribution via $$p(\boldsymbol{\beta} \mid M, \boldsymbol{\theta}) p(M \mid \boldsymbol{\theta}).$$ The update for $p(\boldsymbol{\beta} \mid M, \boldsymbol{\theta})$ is as usual.

## Update for model

The conditional Bayes factor
\begin{align*}
 p(M \mid \boldsymbol{Y}, \boldsymbol{\theta}) &\stackrel{M}{\propto} p(M) p(\boldsymbol{Y} \mid M, \boldsymbol{\theta})
 \\&= p(M) \int_{\mathbb{R}^{\mathbb{|M|}}}p(\boldsymbol{Y} \mid M, \boldsymbol{\beta},\boldsymbol{\theta}) p(\boldsymbol{\beta} \mid M, \boldsymbol{\theta}) d \boldsymbol{\beta}
\end{align*}

## Marginalization

We can  thus marginalize over $\boldsymbol{\beta}$ to get
\begin{align*}
p(\boldsymbol{M} \mid \boldsymbol{Y}, \boldsymbol{\theta}) \propto p(M) |\boldsymbol{Q}_{\boldsymbol{\beta}}|^{-1/2}\exp\left( \frac{1}{2} \boldsymbol{\mu}_{\boldsymbol{\beta}}^\top\boldsymbol{Q}_{\boldsymbol{\beta}} \boldsymbol{\mu}_{\boldsymbol{\beta}}\right)
\end{align*}
where $\boldsymbol{\mu}_{\boldsymbol{\beta}}$ and $\boldsymbol{Q}_{\boldsymbol{\beta}}$ are the mean and precision of $p(\boldsymbol{\beta} \mid \boldsymbol{Y}, M, \boldsymbol{\theta}).$

## Moves for variable selection

We consider different types of move for the $k_{\max}$ potential covariates (including interactions, etc.) [@Holmes:2002]

- birth: adding an unused covariate chosen at random from the remaining ones
- death: removing one covariate at random from the current matrix
- swap an active covariate for an unused one.

Only the last type of move preserves the dimension.

## Jacobians for reversible jump

For most moves $J=1$ in this case, except in four cases where the dimension $|M| \in\{1, 2, k_{\max}-1, k_{\max}\}$ and

- $J=2/3$ if $|M|=1$ and we try to add a covariate, or if $|M|=k_{\max}$ and we try to remove a covariate
- $J=3/2$ if $|M|=2$ and we try to remove a covariate, or if $|M|=k_{\max}-1$ and we try to add the last covariate.

## Posterior weights

We can keep track of which variables are active at each iteration of the MCMC and obtain the marginal posterior probability of inclusion through sample proportions.

This methods that explores neighbouring models (Grey code) only works with a limited number of covariates $p < 25.$

## References

