---
title: "Lecture 4"
author: "Léo Belzile"
toc: false
---

# Rejection sampling


## Beta density with Gaussian proposal

Consider a symmetric beta distribution  and suppose we don't know how to generate from it for a second... Since the mean is at $0.5$, we could propose  $X \sim \mathsf{Gauss}(0.5, \sigma)$ or a truncated version thereof on [$0,1$].

We can compute the value of $C^{\star} = \mathsf{argmin}_\sigma\mathsf{argmax}_xp(x)/q(x; \sigma)$ or equivalently of the difference of log densities, since this is more numerically stable. We do so below using numerical optimization.

```{r}
logpx <- function(x){ dbeta(x, shape1 = 3, shape2 = 3, log = TRUE)}
logqx <- function(x, sd0){ dnorm(x, mean = 0.5, sd = sd0, log = TRUE)}
# Find the worst case C for the Gaussian with sd = sd0
opt_C <- function(sd0){
  logC <- optimize(f = function(x){logpx(x) - logqx(x, sd0)}, 
         maximum = TRUE, 
         lower = 0, 
         upper = 1)$objective
  exp(logC)
}
C_opt <-  optimize(f = opt_C, lower = 0, upper = 100)
# Best fitting standard deviation
(sd0 <- C_opt$minimum)
(C <- C_opt$objective)
```

We can compare the plot and the scaled proposal density, and see how
```{r}
# Plot the curve and the proposal
curve(
  expr = dbeta(x, 3,3), 
  from = 0, 
  to = 1, 
  ylab = "density",
  ylim = c(0, 2.5))
curve(dnorm(
  x, 0.5, sd = sd0)*C, 
  from = 0, 
  to = 1, 
  col = 2, 
  lwd = 2, 
  add = TRUE)
```

We could reduce $C$ further by truncating the Gaussian to the unit interval, as some proposals fall outside. This probability, with the optimal value of $\sigma_0$, is quite small.
```{r}
pnorm(q = 1, mean = 0.5, sd = sd0) - pnorm(q = 0, mean = 0.5, sd = sd0)
```
All draws larger than 1 or smaller than 0 have probability zero
since they are outside of the support of the beta distribution. We use rejection sampling and compute the acceptance rate, comparing the empirical with the theoretical value of 1/C


```{r}
B <- 1e4L # number of proposals
prop_qx <- rnorm(B, mean = 0.5, sd = sd0)
keep <- -rexp(B) < logpx(prop_qx) - logqx(prop_qx, sd = sd0) - log(C)
samp <- prop_qx[keep]
mean(keep) # empirical acceptance rate
1 / C #theoretical acceptance rate
```
Finally, we can check through standard methods that the samples we obtain indeed follow a beta distribution, below using a Kolmogorov-Smirnov test and looking at the quantile-quantile plot.

```{r}
# Quantile-quantile plot
plot(x = qbeta(ppoints(length(samp)), shape1 = 3, shape2 = 3),
     y = sort(samp),
     panel.first = {abline(a = 0, b = 1)},
     xlab = "theoretical quantiles",
     ylab = "empirical quantiles")
# Test of equality of distribution
ks.test(x = samp, pbeta, shape1 = 3, shape2 = 3)
```

## Marsaglia algorithm for Gaussian upper tail

We consider next the Marsaglia algorithm to simulate from the upper tail of a standard normal distribution.

```{r}
# The usual way does not work
a0 <- 10
pa <- pnorm(-a0, lower.tail = FALSE)
qnorm(pa + runif(1) * pa)


a_seq <- seq(10, 50, by = 10)
# The usual way does not work due to numerical overflow
pnorm(a_seq)
```

```{r}

# Marsaglia algorithm
a <- 50
niter <- 1000L
X <- sqrt(a^2 + 2*rexp(niter))
samp <- X[runif(niter)*X <= a]
```

```{r}
# Compute reciprocal of bound C
recipC <- sapply(a_seq, function(a){
  integrate(f = function(x){
    a * exp(-x^2/2 + a^2/2)}, 
    lower = a, 
    upper = Inf)$value})
1/recipC
```


# Importance sampling

Here, we showcase the use of a rather exotic mixture model to reduce the variance of the Monte Carlo estimator of the variance of a symmetric beta distribution.

```{r}
B <- 2e4L; alpha <- 1.5; factor <- 3
# Mode at the mean 0.5
X0 <- rbeta(n = B, alpha, alpha)
px <- function(x){dbeta(x, alpha, alpha)}
# Importance sampling density - mixture of two betas (alpha, factor*alpha)
X1 <- ifelse(
  runif(B) < 0.5,
  rbeta(B, alpha, factor*alpha), 
  rbeta(B, factor*alpha, alpha))
qx <- function(x){
  0.5*dbeta(x, alpha, factor*alpha) + 0.5*dbeta(x, factor*alpha, alpha)}
# Function to integrate - gives variance of a symmetric beta distribution
g <- function(x){(x - 0.5)^2}
# Weights for importance sampling
w <- px(X1)/qx(X1)
```

```{r}
# Monte Carlo integration
mc_est <- mean(g(X0))
mc_var <- var(g(X0))/B
# Importance sampling weighted mean and variance
is_est <- weighted.mean(g(X1), w = w) # equivalent to mean(g(X1)*w)/mean(w)
is_var <- sum(w^2*(g(X1) - is_est)^2)/ (sum(w)^2)
# True value for the beta variance
th_est <- 1/(4*(2*alpha+1))
# Point estimates and differences
round(c(true = th_est, 
        "monte carlo" = mc_est, 
        "importance sampling" = is_est),
      4)
# Ratio of std. errors for means
mc_var/is_var # value > 1 means that IS is more efficient
```




# Estimation of the standard error

We consider a simple $\mathsf{AR}(1)$ process with $\phi=0.8$ and $\sigma=2.$ The latter could be simulated recursively, but we opt her for standard **R** functions.

```{r}
B <- 1e4L
sigma <- 2
phi <- 0.8
x <- arima.sim(
  model = list(ar = phi), 
  n = B,
  rand.gen = rnorm, 
  sd = sigma) # std. deviation of innovations
var(x) 
```
We notice that the estimated variance is not $\sigma^2$! The estimated sample variance equals (up to Monte Carlo precision) the theoretical value of the unconditional variance for the $\mathsf{AR}(1)$, namely $\sigma^2/(1-\phi^2).$

The autocorrelation at lag $h$ is $\phi^h.$ We can compare the sample correlogram with the theoretical value. There is some noise for large lags, due in part to the smaller sample size. Here, since $B$ is very large, everything is well estimated.

```{r}
acf_x <- acf(x, main = "First-order autoregressive")
points(0:30, 0.8^(0:30), type = "p", col = 2)
```

We can get the theoretical effective sample size (ESS) from the formula for the unconditional variance, and by calculating the geometric series. From this, we can also retrieve the standard error for the sample mean of the $B$ draws:

```{r}
ess_factor_ar1 <- 1 / (-1 + 2 / (1 - phi))
uncond_var_ar1 <- sigma^2 / (1 - phi^2)
# Compare with var(x)
se_mean_ar1 <- sqrt(uncond_var_ar1 / B / ess_factor_ar1)
```
 
Below, we consider calculations of the standard error of the sample mean, See the [following simulation for a numerical comparison](https://avehtari.github.io/rhat_ess/ess_comparison.html).

The first method uses the overlapping batch means, which is not recommended. Here, we build groups of size 40, which is sufficient since the autocorrelation tails of quickly.
```{r}
sqrt(c(mcmc::olbm(x, batch.length = niter / 40)))
```
Overlapping batch means are **not** recommended. Rather, Geyer (1992) recommends using the initial sequence method on the batch means output.
```{r}
# Initial convex sequence of Geyer (1992)
geyer_con_var <- mcmc::initseq(x)$var.con
se2 <- sqrt(geyer_con_var / B)
ess2 <- var(x) / geyer_con_var * B
# Also posterior::ess_basic() for Stan output
posterior::ess_basic(matrix(x, ncol = 1))
```

The other estimators are based on spectral analysis, due to Heidelberg and Welch (1981). We estimate the spectrum at frequency zero to get the variance (with the adjustment factor and scale the latter by $B$ to get the variance of the sample mean.

We show below the result of fitting the parametric $\mathsf{AR}(p)$ approximation and using the spectra of the latter; the proposal from Heidelberg and Welch is implemented in `coda::spectrum0`.

```{r}
ar_mod <- ar.yw(x)
# Compute the unconditional variance of the AR process
# # $var.pred is the variance of the innovations
# Formula for the spectrum of AR(p) at 0
#  see, e.g., Percival and Walden, 2020, eq. 446b
uvar <- ar_mod$var.pred / (1 - sum(ar_mod$ar))^2
se3 <- sqrt(uvar / B) # variance of sample mean of AR(p)
ess3 <- var(x) / uvar * length(x)
# directly via coda::effectiveSize
# The methodology from Heidelberg and Welch rather fits a
# gamma GLM to the log periodogram with a polynomial trend
coda::spectrum0(x)
```
