---
title: "Software and packages"
---

There are numerous options for Bayesian scientific computing, including probabilistic programming languages. I will be using [Stan](https://mc-stan.org/), a probabilistic programming language using Hamiltonian Monte Carlo, as it offers cross-platform support.  [JAGS](https://mcmc-jags.sourceforge.io/) has very good automatic differentiation, but is more old-school Gibbs sampling. Other interfaces are platform-specific: below is a list of the main software (**R**, Python and Julia), along with associated packages.


# **R**

The **R** programming language contains many packages on the Comprehensive R Archive Network (CRAN) dedicated for Bayesian analysis; see the [CRAN Task View](https://cran.r-project.org/web/views/Bayesian.html) on Bayesian statistics.

Among the packages worth exploring are `NIMBLE` (<https://r-nimble.org/>) and [Greta](https://greta-dev.github.io/greta/index.html).

To install Stan with **R**, use the following commands:

```{r}
#| eval: false
#| echo: true
install.packages("cmdstanr",
                 repos = c("https://mc-stan.org/r-packages/",
                           getOption("repos")))
```

I also use `priorsense`, `bayesplot` and `loo` packages for day-to-day analysis. Good high-level interface to Stan in **R** include `brms`, which allows for regression models with formula interface.

For beginners (without knowledge of Markov chain Monte Carlo), the packages `bang` + `rust`, and [`rethinking`](https://github.com/rmcelreath/rethinking) from  @McElreath:2020 are also useful.

- The `INLA` **R** package (<https://www.r-inla.org/>), which uses integrated nested Laplace approximations (INLA) for inference in latent Gaussian models, is not on the CRAN.

```{r}
#| eval: false
install.packages("BiocManager")
BiocManager::install("graph")
BiocManager::install("Rgraphviz")
install.packages(
  "INLA",
  repos = c(getOption("repos"),
            INLA = "https://inla.r-inla-download.org/R/stable"),
  dep = TRUE)
```

I have bundled many datasets used in class in a **R** package available from this Github repository:

```{r}
#| eval: false
# Download databases for course
remotes::install_github("lbelzile/hecbayes")
```


# Python

Python can be installed through the [`uv` package manager](https://docs.astral.sh/uv/), which I recommend over other alternatives. 

Beyond Stan via `CmdStanPy`, the most popular Python library for Bayesian data analysis is [`PyMC`](https://www.pymc.io/welcome.html).


# Julia

The [Julia Programming Language](https://julialang.org/) (<https://julialang.org/>) has [numerous packages](https://juliapackages.com/c/bayesian) for Bayesian analysis,
including notably

- [`Pigeons`](https://pigeons.run/dev/) 
- [`Turing`](https://turinglang.org/)
