---
title: "Solution 4"
---


## Exercise 4.1

Consider the Laplace family of distribution, $\mathsf{Laplace}(\nu, \tau)$, with density
\begin{align*}
g(x; \nu, \tau) = \frac{1}{2\tau} \exp\left(- \frac{|x-\nu|}{\tau}\right), \qquad \nu \in \mathbb{R}, \tau > 0
\end{align*}
as a candidate distribution for rejection sampling from $\mathsf{Gauss}(0,1)$.

1. Provide an inversion sampling algorithm to generate from $\mathsf{Laplace}(\nu, \tau)$.
2. Can you use the proposal to generate from a standard Gaussian? for Student-$t$ with 1 degree of freedom? Justify your answer.
3. Consider as proposal a location-scale version of the Student-t with $3$ degrees of freedom. Find the optimal location and scale parameters and the upper bound $C$ for your choice.
4. Use the accept-reject to simulate 1000 independent observations and compute the empirical acceptance rate.



::: {.solution}

1. The distribution function is 
\begin{align*}
F(x) = \begin{cases}
\frac{1}{2} \exp\left(\frac{x-\nu}{\tau}\right) & x \leq \nu\\
1 - \frac{1}{2} \exp\left(-\frac{x-\nu}{\tau}\right) & x >\nu\\
\end{cases}
\end{align*}
and using the quantile transform, set $X=\nu + \tau \log(2U)$ if $U \leq 0.5$ and $X=\nu - \tau\log(2-2U)$ if $U > 0.5$ for $U \sim \mathsf{unif}(0,1)$.
2. The Gaussian has lighter tail than the Laplace, so this won't work. The Cauchy distribution would be a suitable candidate, albeit too heavy tailed.
3. The optimal value for the location of the Student-$t$ would be $\nu$ (e.g., zero for the standard Laplace). We compute the optimal scale via in the code below:
$$
\mathrm{argmin}_{\sigma \in \mathbb{R}_{+}}\mathrm{argmax}_{x \in \mathbb{R}} \{\log f(x) - \log g(x; \sigma)\}
$$


```{r}

#' Laplace density
dlaplace <- function(x, loc = 0, scale = 1, log = FALSE){
 stopifnot(scale > 0)
 logdens <-  -log(2*scale) - abs(x-loc)/scale
 if(log){
 return(logdens)
 } else{
 return(exp(logdens))
 }
}
dstudent <- function(x, loc = 0, scale = 1, df = 1, log = FALSE){
  logdens <- -log(scale) + dt(x = (x - loc)/scale, df = df, log = TRUE)
   if(log){
    return(logdens)
   } else{
   return(exp(logdens))
   }
}
# For each value of the scale sigma,
# find the minimum value of x (typically at zero)
opt <- optimize(f = function(sigma){
  optimize(f = function(x){
  dlaplace(x, log = TRUE) - 
    dstudent(x, scale = sigma, df = 3, log = TRUE)}, 
  maximum = TRUE, 
  interval = c(-100, 100))$objective
}, interval = c(0.1,10))
(C <- exp(opt$objective))
(sigma <- opt$minimum)

# Simulate from accept-reject
ntrials <- 1.1*C*1000
# Simulate from location-scale student
candidate <- sigma*rt(n = ntrials, df = 3)
# Compute log of acceptance rate
logR <- dlaplace(candidate, log = TRUE) - 
  dstudent(candidate, scale = sigma, df = 3, log = TRUE) 
samp <- candidate[logR >= log(C) + -rexp(ntrials)]
# Monte Carlo estimator of the acceptance rate
ntrials/length(samp)
# Plot density
library(ggplot2)
ggplot(data = data.frame(x = samp),
       mapping = aes(x = x)) +
  geom_density() +
  theme_classic()
```
4. The Monte Carlo acceptance rate is `r round(ntrials/length(samp), 2)`, compared with the analytical bound found via numerical optimization of `r round(C, 2)`, to two significant digits.

:::


## Exercise 4.2

We revisit [Exercise 2.3](/exercises/02-exercise.html#exercice-2.3), which used a half-Cauchy prior for the exponential waiting time of buses.

The ratio-of-uniform method, implemented in the [`rust` **R** package](https://paulnorthrop.github.io/rust/index.html), can be used to simulate independent draws from the posterior of the rate $\lambda$.
The following code produces 
```{r}
#| eval: true
#| echo: true
nobs <- 10L # number of observations
ybar <- 8   # average waiting time
B <- 1e4L  # number of draws
# Un-normalized log posterior: scaled log likelihood + log prior
upost <- function(x){ 
  dgamma(x = x, shape = nobs + 1L, rate = nobs*ybar, log = TRUE) +
    log(2) + dt(x = x, df = 1, log = TRUE)}
post_samp <- rust::ru(logf = upost, 
                      n = B, 
                      d = 1,  # dimension of parameter (scalar)
                      init = nobs/ybar)$sim_vals # initial value of mode
```

Estimate using the independent Monte Carlo samples:

1. the probability that the average waiting time $1/\lambda$ is between 3 and 15 minutes
2. the average waiting time
3. the standard deviation of the average waiting time.


Next, implement a random walk Metropolis--Hastings algorithm to sample draws from the posterior and re-estimate the quantities. Compare the values.

::: {.solution}

```{r}
# Monte Carlo for waiting time
# Lambda is the reciprocal mean (1/minute)
times_mc <- 1/post_samp
# Mean of binary variables + std. error
p <- mean(times_mc > 3 & times_mc < 15)
c(est = p, se = sqrt(p*(1-p)/B))
# Posterior mean with standard error
c(est = mean(times_mc), se = sd(times_mc)/sqrt(length(post_samp)))



# Metropolis-Hastings algorithm
B <- 1e4L
curr <- 8/10 # prior mean
chains <- numeric(B) # container
sd_prop <- 0.1 # proposal standard deviation
for(b in seq_len(B)){
  prop <- rnorm(n = 1, mean = curr, sd = sd_prop)
  if(upost(prop) - upost(curr) > -rexp(1)){
    curr <- prop
  }
  chains[b] <- curr
}
# Discard burn-in
times <- 1/chains[-(1:100)]
# Estimate quantities using Monte Carlo
mean(times > 3 & times < 15)
mean(times)
sd(times)
# Summary of MCMC
mcmc <- coda::mcmc(times)
summary(mcmc)
```
We can see that the standard error for the mean is roughly twice as big. We would thus need to inflate the sample size by a factor four to get the same precision.

:::
