---
title: "Solution 7"
---
  
```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
library(brms)
library(INLA)
library(patchwork)
library(ggplot2)
theme_set(theme_classic())
```


## Exercise 7.1

Consider the eight school data from Educational Testing Service (ETS), featured prominently in @Gelman:2013. The data shows results of randomized experiments in eight different schools to estimate the effect of coaching on SAT-V scores.
The data consist of estimated treatment mean difference and their standard errors. The data size are large, so the average treatment effect (ATE) can be considered Gaussian and the standard errors are treated as a known quantity.



1. Derive the Gibbs sampler for the two models with improper priors. Note that, to sample jointly from $p(\mu, \boldsymbol{\alpha} \mid \boldsymbol{y}, \sigma^2_{\alpha}, \boldsymbol{\sigma}^2_{\boldsymbol{y}})$, you can draw from the marginal of $\boldsymbol{\alpha}$, then the conditional $\mu \mid \boldsymbol{\alpha}, \cdots$.
2. Fit the three algorithms described in @Gelman.vanDyk:2008 and compare their efficiency via effective sample size (ESS). In particular, draw traceplots and calculate effective sample sizes for 
    
    - $\mu$
    - $\alpha_1$, 
    - the sum $\mu + \alpha_1$.

as well as pairs plot of $\alpha_i, \log \sigma_{\alpha}$. The latter should display a very strong funnel.



```{r}
#| eval: true
#| echo: false
library(ggplot2)
es <- data.frame(school = LETTERS[1:8], 
                 y = c(28, 8, -3, 7, -1, 1, 18, 12), 
                 se = c(15, 10, 16, 11, 9, 11, 10, 18))
ggplot(es, aes(x = school, y = y)) +
  geom_point(col = "purple") +
  geom_errorbar(aes(ymin = y - 1.96*se, ymax = y + 1.96*se), width = 0.3) +
  labs(x = "school", y = "",
       subtitle = "Average treatment effect with asymptotic 95% confidence intervals")
```


Consider a one-way ANOVA with a single observation per group, where for $i=1, \ldots, n,$ we have
\begin{align*}
Y_{i} &\sim \mathsf{Gauss}(\mu + \alpha_i, \sigma^2_i) \\
\alpha_i &\sim \mathsf{Gauss}(0, \sigma^2_\alpha)
\end{align*}
and an improper prior for the mean $p(\mu) \propto 1.$
If there are $K$ groups, then the group-specific mean is $\mu+\alpha_k$, and there is a redundant parameter. In the Bayesian setting, the parameters are weakly identifiable because of the prior on $\boldsymbol{\alpha}$, but the geometry is such that $\alpha_k \to 0$ as $\sigma^2_{\alpha} \to 0.$ See [Section 5.5](https://users.aalto.fi/~ave/BDA3.pdf) on p. 119 of @Gelman:2013 for more details.

We fit the model by adding redundant parameter to improve mixing [@Gelman.vanDyk:2008]
\begin{align*}
Y_{i} &\sim \mathsf{Gauss}(\mu + \alpha\xi_i, \sigma^2_i) \\
\xi_i &\sim \mathsf{Gauss}(0, \sigma^2_\xi)
\end{align*}
so that $\sigma_\alpha = |\alpha|\sigma_\xi.$


See this [post](https://xuwd11.github.io/am207/wiki/gelmanschools.html) for display of the pathological behaviour that can be captured by divergence in Hamiltonian Monte Carlo.


:::{.solution}

```{r}
#| eval: true
#| echo: true

# Eight school data
es <- data.frame(school = LETTERS[1:8], 
                 y = c(28, 8, -3, 7, -1, 1, 18, 12), 
                 se = c(15, 10, 16, 11, 9, 11, 10, 18))
n <- nrow(es)
B <- 1e3L
set.seed(80601)
# Starting values
alpha <- rnorm(0, sd = 0.01)
mu <- mean(es$y)
sigma_al <- sqrt(var(es$y) - mean(es$se))
pars1 <- matrix(NA, nrow = B, ncol = n + 2)
var_mu <- 1/sum(1/es$se^2)
for(b in seq_len(B)){
  var_al <- 1/(1/es$se^2 + 1/sigma_al^2)
  pars1[b, 1:n] <- alpha <- rnorm(n = n, mean = var_al * (es$y - mu)/es$se^2, sd = sqrt(var_al))
  pars1[b, n + 1] <- mu <- rnorm(n = 1, mean = var_mu * sum((es$y - alpha)/es$se^2), sd = sqrt(var_mu)) 
  pars1[b, n + 2] <- sigma_al <- sum(alpha^2) / rchisq(1, df = n - 1) 
}

## Sampler 2
# Same model, with joint updates
set.seed(80601)
# Starting values
alpha <- rnorm(0, sd = 0.01)
mu <- mean(es$y)
sigma_al <- sqrt(var(es$y) - mean(es$se))
pars2 <- matrix(NA, nrow = B, ncol = n + 2)
for(b in seq_len(B)){
  var_al <- 1/(1/es$se^2 + 1/sigma_al^2)
  # Sample from joint of alpha, mu
  # by taking p(alpha | sigma_al, y) * p(mu | alpha, sigma_al, y)
  pars2[b, 1:n] <- alpha <- rnorm(n = n, mean = var_al * es$y/es$se^2, sd = sqrt(var_al))
  pars2[b, n + 1] <- mu <- rnorm(n = 1, mean = var_mu * sum((es$y - alpha)/es$se^2), sd = sqrt(var_mu)) 
  pars2[b, n + 2] <- sigma_al <- sum(alpha^2) / rchisq(1, df = n - 1) 
}

# Parameter expansion - V+PX sampler
a <- 1
pars3 <- matrix(NA, nrow = B, ncol = n + 3)
for(b in seq_len(B)){
  var_al <- 1/(1/es$se^2 + 1/sigma_al^2)
  # Sample from joint of alpha, mu
  # by taking p(alpha | sigma_al, y) * p(mu | alpha, sigma_al, y)
  alpha_st <- rnorm(n = n, mean = var_al * es$y/es$se^2, sd = sqrt(var_al))
  pars3[b, n + 1] <- mu <- rnorm(n = 1, mean = var_mu * sum((es$y - alpha)/es$se^2), sd = sqrt(var_mu)) 
  pars3[b, n + 2] <- sigma_al_st <- sum(alpha_st^2) / rchisq(1, df = n - 1) 
  pars3[b, n + 3] <- a <- rnorm(n = 1, mean = sum(alpha_st*(es$y-mu)/es$se^2)/sum(alpha_st^2/es$se^2), 
                                sd = sqrt(sum(alpha_st^2/es$se^2)))
  pars3[b, 1:n] <- alpha_st*a
  pars3[b, n + 2] <- sigma_al <- abs(a)*sigma_al_st
}
```

We can write the joint density of $\mu$ and $\boldsymbol{\alpha}$ and integrate out $\mu$. The precision for $\alpha_j$ is $Q_{\alpha_j} = 1/\sigma^2_j + 1/\sigma_{\alpha}^2$, and the unconditional mean of $p(\alpha_j \mid \sigma_{\alpha})$ is $Q^{-1}_{\alpha_j}y_j/\sigma^2_j$. The conditional distribution $p(\mu, \mid \boldsymbol{\alpha}, \sigma_{\alpha})$ is Gaussian with precision $Q_{\mu} = \sum_{j=1}^{8} \sigma^{-2}_j$ and mean $Q_{\mu}^{-1}\sum_{j=1}^8 \sigma^{-2}_j(y_j-\alpha_j).$ The other steps are detailed in @Gelman.vanDyk:2008.

```{r}
#| eval: true
#| echo: false
# Compare trace plots for mu + alpha
plot(pars1[, 2] + pars1[, n+1])
plot(pars1[, 2])
# Funnel behaviour
g1 <- ggplot(
  data = data.frame(
   beta2 = c(pars1[,2]), 
   sigma = c(pars1[,n+2])),
  mapping = aes(x = beta2, y = log(sigma))) +
  geom_point() +
  scale_color_grey() +
  labs(x = expression(alpha[2]), 
       y = expression("log"~sigma[alpha]),
       caption = "block") 
g2 <- ggplot(
  data = data.frame(
   beta2 = c(pars3[,2]), 
   sigma = c(pars3[,n+2])),
  mapping = aes(x = beta2, y = log(sigma))) +
  geom_point() +
  scale_color_grey() +
  labs(x = expression(alpha[2]), 
       y = expression("log"~sigma[alpha]),
       caption = "expansion") 
g1 + g2
```

:::
